### 2024-12-03 paper/paper.qmd
soft reset

---
title: "Younger age groups, women, and lower levels of education are more likely to vote Democractic"
subtitle: "A logistic regression analysis of demographics in the 2020 US Presidential Election"
author: 
  - Victor Ma
thanks: "Code and data are available at: https://github.com/bestmustard/Determinants-of-Political-Support-in-the-United-States."
date: 03/16/2024
date-format: long
abstract: "In this study, I investigate demographic patterns which influence political support for the US 2020 election, specifically focusing on whether demographics of people tend to vote for Joe Biden or Donald Trump. I use a logistic regression model with the variables age, gender, and education to predict which categories in each demographic have the highest chance of voting for Joe Biden. I find that the model accurately predicts the trend of younger people, women, and those with the lowest level of education being more likely to support Joe Biden."
format: pdf
number-sections: true
bibliography: references.bib
editor: 
  markdown: 
    wrap: 72

header-includes:
 \usepackage{float}
 \floatplacement{figure}{H}
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(dataverse)
library(knitr)
library(modelsummary)
library(rstanarm)
library(tidybayes)
library(dplyr)
library(patchwork)
library(ggplot2)
library(broom.mixed)
ces_data <- read_csv("../data/analysis_data/analysis_data.csv")
ces_data$voted_for <- as_factor(ces_data$voted_for)
ces_data$education <- factor(ces_data$education, levels = c("No HS", "High school graduate", "Some college", "2-year", "4-year", "Post-grad"))
```

# Introduction

The 2020 United States presidential election was not only a pivotal political event but also a reflection of a nation grappling with unprecedented challenges. The backdrop of the COVID-19 pandemic, which affected millions and significantly altered everyday life, coupled with nationwide protests following the death of George Floyd, underscored the deep social and political divisions within the country [@KFF2020; @Cohn2020]. These events created a heightened sense of importance around the election that had potential to shape the direction of national policies. 2020 was a year of heightened misinformation and polarization, with many people challenging the electorate's ability to navigate complex issues and make informed decisions [@Lazer2018; @Allcott2017].

Understanding how demographics are correlated with political alignment in such a pivotal year reveals a lot about societal trends. For example, younger voters might prioritize progressive issues like climate change and social justice, signal a shift towards liberalism. Older voters, emphasizing economic and healthcare concerns, show more conservative preferences. Gender disparities in voting reveal women's tilt towards equality-focused candidates, while varying education levels demonstrate that higher education correlates with liberal views, highlighting the role of informed decision-making in political orientation. This demographic analysis not only maps the political landscape of 2020 but also teaches us about evolving societal values.

In this study I use a logistic regression model in order to predict which candidate will win more votes based on demographics such as age, gender, and level of education. The results of this model gives us key insights into how much demographic variables can influence voter preferences in a polarized climate. In particular, logistic regression models are used to model binary outcomes, so our outcome will be whether Joe Biden or Donald Trump is forecasted to have more votes.

The data we are using is the 2020 Cooperative Election Study (CCES), an American stratified sample survey administered by YouGov [@ces2020]. The data is directly sourced from the Harvard dataverse, an online data repository running on the open-source web application dataverse. For my purposes I will be focusing on the variables of age, gender, and level of education as explanatory variables in the logistic regression model. I will also be referencing data from similar polls collected for the 2020 election, such as from American National Election Studies, Gallup Polls, and Pew Research Surveys [@ANES2020; @Pew2020; @Gallup2020]

My report is structured into four main sections following the introduction. In the first section, I describe the data utilized for my analysis, highlighting the CCES 2020 dataset and presenting graphs that show the distribution of key demographic variables. The second section details the logistic regression model, including the rationale for its use and an interpretation of preliminary findings. Next I will analyse the demographic factors' impact on election outcomes through the use of graphs and specific numerics from my results. Finally, I discuss the implications of my findings, address potential weaknesses in my study, and suggest directions for future research.

This analysis is conducted using R, using several R packages to facilitate my analysis and presentation. This includes tidyverse for data manipulation and visualization, dataverse for accessing the CCES data, knitr for report generation, modelsummary for model interpretation, and rstanarm for Bayesian regression modeling [@citeR; @Dataverse; @Knitr; @Modelsummary; @Rstanarm; @Tidybayes; @Dplyr]. Some portions including ggplot graphs and the "Discussion" section were written with the help of ChatGPT4 @chatgpt. 
\newpage

# Data
The dataset I used was the 2020 Cooperative Election Study (CCES) [@ces2020]. This data is provided by the Harvard dataverse and is conducted yearly as a survey of US political opinions, with the 2020 iteration including 61,000 respondents @ces2020. The release also includes a full guide to the data, and the questionnaires used. Vote validation was conducted by Catalist, a largescale organization with information on over 240 million individuals in the United States [@Catalist2017].

## Strengths

### Sample Size

Sources for @fig-respondents below: [@ANES2020],[@Pew2020],[@Gallup2020], [@ces2020].

```{r out.height="30%"}
#| label: fig-respondents
#| fig-cap: "Number of respondents for various 2020 election survey sources"
#| echo: false
#| warning: false
#| message: false

# Create a data frame with the number of respondents for each source
survey_data <- data.frame(
  Source = c("ANES 2020", "Pew Research", "Gallup Polls", "CCES 2020"),
  Respondents = c(8280, 11818, 1000, 61000)
)

# Generate the point graph
ggplot(survey_data, aes(x = Source, y = Respondents)) +
  geom_point(size = 5, color = "blue") +
  theme_minimal() +
  labs(
    title = "Number of Respondents per Survey Source",
    x = "Survey Source",
    y = "Respondents"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) # Angle the x-axis text for readability

```

From @fig-respondents, we can see that the CCES poll has the largest respondent pool compared to similar surveys. This ensures that analyses can be sufficiently granular, allowing us to analyze smaller subgroups that a smaller dataset may not represent accurately [@Ansolabehere2020].

### Data Collection Phases
The CCES collected data in two waves, immediately before and after the 2020 election. This helps capture the shifts in voter attitudes and preferences during the critical final stages of the election campaign [@ces2020]. Pew Research and Gallup each conduct multiple polls of much smaller scale leading up to the election, which will give a snapshot in time of the voter sentiment but does not capture as broad of a perspective as the CES survey [@Pew2020], [@Gallup2020].

### Sampling Methodology
The sampling methodology and vote validation protocols are well outlined in the guide. This dataset employs sample matching, a two-step process designed to create a survey sample that represents a larger population as closely as possible. First, a sample that reflects the broader population (called the target sample) is selected based on certain characteristics like age, race, and gender. Then, for each person in this target sample, someone with the closest matching profile from a pool of survey volunteers is chosen. The match is based on many characteristics available in databases, ensuring the survey sample mirrors the target population.

After the matched sample is created, it's given weights to adjust for any minor differences between it and the target population, ensuring the final results are representative. The weights are calculated in two stages using data from the American Community Survey and validated voter registration records, accounting for factors such as demographics and voter behaviors.

The CES then uses state-level samples and compares them with actual election results to validate the sample's accuracy and the weighting process. If survey estimates closely align with the actual votes, it suggests that the CES sample is accurate and representative [@ces2020].

This methodology is well fitted for the online opt-in nature of the survey, and helps combat the issues that arise with random or quota sampling which lead to results that are uncharacteristic of the target sample [@sample2007].

## Limitations
While the data is largely reliable, there exist some smaller errors and limitations. The guide details an error affecting 925 respondents from North Carolina, who were shown candidate names for incorrect districts. This could influence the accuracy of analyses related to House races, necessitating verification methods and adjustments in studies which focus on these aspects.

As with any survey-based research, the CCES 2020 relies on self-reported data, which can introduce biases such as social desirability bias or recall bias. Additionally, pre-election surveys may not capture last-minute shifts in voter sentiment, which is particularly relevant in a rapidly evolving and polarized political climate like that of 2020.

The survey was conducted using an online survey platform hosted by YouGov [@ces2020]. Online surveys overcome geographical and physical barriers that traditional survey methods might face, allowing for more diverse participant engagement. However, certain demographic groups might be underrepresented in online panels due to varying levels of internet access and technological literacy, potentially introducing biases into the dataset. Additionally, the impersonal nature of online surveys could affect the quality of responses, since the lack of a direct interviewer may lead to less thoughtful answers from participants.

## Variables of Interest
The selection of age, gender, and education as predictor variables for analyzing political preference in the 2020 CCES data is informed by specific findings from prior research that shows their impact on voting behavior.

### Age
The impact of age on political preferences is significant, as demonstrated in the 2016 election context. An article by Pew Research (2018) shows that 58% of validated voters aged 18-29 voted for the Democratic Party (Hillary Clinton), compared to 28% voting Republican (Trump). These numbers slowly begin to favor Trump with older age groups, where 53% of validated voters aged 65+ voted for Trump and only 44% voted for Hillary Clinton [@2016votes]. Research by Iyengar and Krupenkin (2018) illustrates that younger voters, influenced by contemporary sociopolitical developments, exhibit a marked preference for progressive policies and candidates [@IyengarKrupenkin2018]. This tendency is often contrasted with older voters who may prioritize different issues based on their lived experiences.

### Gender
The gender gap in political preferences is a well-documented phenomenon, with women more likely to support democratic candidates and policies, in particular social welfare and healthcare [@BoxSteffensmeier2004; @Dolan2014]. Pew Research found 11% more men voted for Trump compared to Clinton, while 16% more women chose the Democratic candidate Clinton [@2016votes].

### Education
Level of education is a strong predictor of political knowledge and ideological orientation. Individuals with higher education levels are generally more politically engaged, exhibit greater political efficacy, and tend to support liberal ideologies [@Highton2009]. We saw this play out in the 2016 election as generally college-grads were more favoured toward the Democratic Party, while remarkably more (64% vs. 28%) non-grads favoured Trump [@2016votes].

## Data Preparation and Cleaning
The 2020 CCES data was first obtained via the dataverse package directly from Harvard Dataverse, and then saved in parquet format using the arrow package for efficient storage and access [@arrow]. The cleaning process involved filtering for registered voters who chose either Biden or Trump, in order to treat presidential votes as a binary outcome.

The respondent variables (age, gender, and education) were transformed for clarity. Age was categorized into groups ("65+", "45-64", "30-44", "18-29"), gender was labeled ("Male", "Female"), and education levels were delineated ("No HS", "High school graduate", "Some college", "2-year", "4-year", "Post-grad"). These steps utilized the dplyr package for data manipulation.

The cleaned dataset was then saved in both CSV and parquet formats.

The distributions for each explanatory variable are illustrated in @fig-education and @fig-agegender below:

```{r out.height="20%"}
#| label: fig-education
#| fig-cap: "Number of respondents for various 2020 election survey sources"
#| echo: false
#| warning: false
#| message: false

ggplot(ces_data, aes(x = education, fill = education)) +
  coord_flip()+
  geom_bar() +
  scale_fill_brewer(palette = "Paired") + # More colors for distinction
  theme_minimal() +
  labs(title = "Respondents per Level of Education",
       y = "Education Level",
       x = "Count") +
  theme(legend.title = element_blank())

```

```{r out.height="20%"}
#| label: fig-agegender
#| fig-cap: "Number of respondents for age group and gender"
#| echo: false
#| warning: false
#| message: false
# visualizing the age group distribution
plot_age <- ggplot(ces_data, aes(x = age)) +
  geom_bar(fill = "#FFA07A") +
  theme_minimal() +
  labs(title = "Respondents per Age Group", x = "Age Group", y = "Count") +
  theme(legend.title = element_blank())

# visualizing the gender distribution
plot_gender <- ggplot(ces_data, aes(x = gender)) +
  geom_bar(fill = "#778899") +
  theme_minimal() +
  labs(title = "Respondents per Gender", x = "Gender", y = "Count")+
  theme(legend.title = element_blank())

plot_age + plot_gender + plot_layout(ncol = 2)
```

@fig-education tells us that 4-year degree graduates are the most well represented in the polls, followed by high school graduates, people with some college, and then 2-year degree graduates. People who have not graduated high school are hardly represented at all.

In @fig-agegender we see 45-64 being the most dominant age group, with much less representation in the 18-29 range. Also, about 20% more females participated in the study than males.

Below are some figures representing who people voted for based on their age and gender, followed by education level and gender.

```{r out.height="20%"}
#| label: fig-agevotes
#| fig-cap: "Proportion of votes at each age group"
#| echo: false
#| warning: false
#| message: false

# Relationship between Age and Political Preferences
ggplot(ces_data, aes(x = age, fill = voted_for)) +
  geom_bar(position = "fill") +
  scale_fill_brewer(palette = "Set1") +
  theme_minimal() +
  labs(title = "Age and Political Preferences", x = "Age Group", y = "Proportion")
```
@fig-agevotes shows a negative correlation between votes for Joe Biden and older age groups.
```{r out.height="20%"}
#| label: fig-gendervotes
#| fig-cap: "Proportion of votes by gender"
#| echo: false
#| warning: false
#| message: false
ggplot(ces_data, aes(x = gender, fill = voted_for)) +
  geom_bar(position = "fill") +
  scale_fill_brewer(palette = "Set1") +
  theme_minimal() +
  labs(title = "Gender and Political Preferences", x = "Age Group", y = "Proportion")
```
We can see that in @fig-gendervotes, more women voted for Joe Biden than men.

```{r out.height="20%"}
#| label: fig-educationvotes
#| fig-cap: "Proportion of votes at each level of education"
#| echo: false
#| warning: false
#| message: false
# Relationship between Education and Political Preferences

ggplot(ces_data, aes(x = education, fill = voted_for)) +
  coord_flip()+
  geom_bar(position = "fill") +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal() +
  labs(title = "Education and Political Preferences", x = "Education Level", y = "Proportion")
```

@fig-educationvotes is more interesting in the sense that the number of voters for Biden generally trends positively with higher levels of education, but the "No HS" and "Some College" groups had more votes for Biden respectively than the next level of education. It may be worth to note that "Some college" does not specify how much time they spent in college.

# Model

Logistic regression is a model used when the outcome or dependent variable is binary, which fits this scenario perfectly as we are modelling votes for Joe Biden vs. Donald Trump.

The regression model will calculate the log odds of the probability that a respondent votes for Joe Biden, and then map it to a probability between 0 and 1 through the logistic function.

The standard logistic function $\sigma(t)$ for a real-valued input $t$ is defined as:

```{=tex}
\[ \sigma(t) = \frac{1}{1 + e^{-t}} \]
```

The graph of the logistic function is an S-shaped curve known as a sigmoid curve. It approaches 1 as $t$ goes to positive infinity and approaches 0 as $t$ goes to negative infinity.

In logistic regression, the input $t$ is the linear combination of predictors including the intercept, which can be represented as $\beta_0 + \beta_1X_1 + \beta_2X_2 + ... + \beta_nX_n$. The logistic function then translates this into a probability that the dependent variable is 1 (voted for Joe Biden).

In this situation, I will be using the predictors age, education, and gender and then applying the logistic function to get the probability $P(Y_i=1)$ that a respondent $i$ supports Joe Biden.

This model is particularly strong at handling categorical dependent variables, which each of my explanatory variables fall under [@Hosmer2013].

## Model Specification

The logistic regression model is defined as:

```{=tex}
\[
\log\left(\frac{P(Y_i=1)}{1 - P(Y_i=1)}\right) = \beta_0 + \beta_1X_{\text{age},i} + \beta_2X_{\text{education},i} + \beta_3X_{\text{gender},i}
\]
```
## Model set-up

- $Y_i$ is the binary indicator of support for the Democratic candidate Joe Biden (1) versus the Republican candidate Donald Trump (0) for respondent $i$.
- $X_{\text{age},i}$, $X_{\text{education},i}$, and $X_{\text{gender},i}$ are the age, education level, and gender of respondent $i$, respectively.
- $\beta_0$ represents the model intercept, while $\beta_1$, $\beta_2$, and $\beta_3$ are coefficients quantifying the effects of age, education, and gender on the likelihood of Democratic support.

I fit my logistic regression model to the data using 'stan_glm()' function from the 'rstanarm' package in R [@citeR, @Rstanarm]. This function will automatically determine each of the $\beta$ coefficients in the model, using a smaller slice sample of 3000 from the 2020 CCES data we processed. This function also uses Bayesian logistic regression with the default priors from 'rstanarm'.

### Model Justification

In order to better interpret the results of the model, I can create a coefficient plot to visually see the effect sizes of the predictor variables on the likelihood of an individual supporting Biden.

@fig-coefficient maps each predictor variable on the y-axis to an effect size and confidence interval on the x-axis. The effect size is the change in log-odds of supporting Biden for a one-unit increase in the predictor variable, which is essentially how much impact each demographic characterstic has an effect on voting for Biden.

The confidence intervals tell us the range within we can be confident that the true effect lies. Smaller confidence intervals means there is a higher level of precision in the estimate of the effect size. I am not as interested in the intervals that cross zero because that means that there is data to support each side (Biden or Trump) and so they are less statistically significant.

```{r}
#| label: fig-coefficient
#| fig-cap: "Coefficient plot of demographics"
#| echo: false
#| warning: false
#| message: false
political_preferences <- readRDS(file ="../models/political_preferences.rds")

# Assuming `tidy_political_preferences` is a tidy version of your model with terms and estimate
tidy_political_preferences <- tidy(political_preferences)

ggplot(tidy_political_preferences, aes(x = term, y = estimate)) +
  geom_point(aes(color = estimate > 0), size = 3) +  # color points based on positive or negative coefficient
  geom_errorbar(aes(ymin = estimate - std.error, ymax = estimate + std.error), width = 0.2) +  #error bars
  coord_flip() +  
  scale_color_manual(values = c("blue", "red")) +  # Set colors for positive and negative coefficients
  labs(title = "Joe Biden Support",
       x = "Predictor Variables",
       y = "Effect on Log Odds of Supporting Biden",
       color = "Effect Direction") +
  theme_minimal()  # Use a minimal theme

# Ensure the 'tidy_political_preferences' contains the columns you want to summarize
tidy_political_preferences <- broom::tidy(political_preferences)

# Check available columns
names(tidy_political_preferences)
```

The conclusions I can draw from @fig-coefficient align with what I expected given the dataset. It's clear that higher education levels, specifically 4-year degrees and post-grad degrees, are positively associated with support for Biden. Higher age groups, specifically 45-64 and 65+, are strongly tied to voting for Trump. Men are less likely to vote for Biden than Women.

\newpage
# Results

Figures @fig-page, @fig-pgender, and @fig-pedu are recreations of the earlier graphs we saw showcasing the proportion of voters for Joe Biden by groups within each demographic, where the bar is the data from CCES 2020 while the point is the prediction generated by the model.

```{r out.height="25%"}
#| label: fig-page
#| fig-cap: "Model prediction for votes by age vs. CCES 2020 data"
#| echo: false
#| warning: false
#| message: false
# Example constant values for the other factors
constant_education <- "2-year" # Choose a representative value
constant_gender <- "Male" # Choose a representative value

# Create a data frame for age predictions
age_data <- data.frame(age = c("18-29", "30-44", "45-64", "65+"),
                       education = constant_education,
                       gender = constant_gender)

# Predict probabilities for each age group
age_data$predicted_prob <- predict(political_preferences, newdata = age_data, type = "response")
library(ggplot2)
# Actual proportions from the data
ces_data_grouped <- ces_data %>%
  group_by(age) %>%
  summarize(Proportion_for_Biden = mean(voted_for == "Biden", na.rm = TRUE))

# Assuming 'political_preferences' is your logistic regression model object and
# 'age_data' contains the predicted probabilities with 'age' as a factor
age_data$age <- factor(age_data$age, levels = c("18-29", "30-44", "45-64", "65+"))

# Create a combined plot
ggplot() +
  geom_bar(data = ces_data_grouped, aes(x = age, y = Proportion_for_Biden, fill = age), 
           stat = "identity", position = "dodge", alpha = 0.5) +
  geom_point(data = age_data, aes(x = age, y = predicted_prob, color = age), 
             position = position_dodge(width = 0.9), size = 3) +
  scale_fill_brewer(palette = "Set1") + 
  scale_color_brewer(palette = "Set1") + 
  labs(title = "Actual vs. Predicted Support for Biden by Age Group",
       x = "Age Group", y = "Proportion") +
  theme_minimal() +
  theme(legend.position = "none")

```

```{r out.height="25%"}
#| label: fig-pgender
#| fig-cap: "Model prediction for votes by gender vs. CCES2020 data"
#| echo: false
#| warning: false
#| message: false
# Example constant values for the other factors (choose representative values for each)
constant_age <- "30-44" # Assuming this is a middle category for age

# Create a data frame for gender predictions
gender_data <- data.frame(
  gender = c("Male", "Female"),
  education = constant_education, # Assuming you have a constant value set for education
  age = constant_age
)

# Predict probabilities for each gender category
gender_data$predicted_prob <- predict(political_preferences, newdata = gender_data, type = "response")

# Create a data frame for education predictions
education_data <- data.frame(
  education = c("No HS", "High school graduate", "Some college", "2-year", "4-year", "Post-grad"),
  gender = constant_gender, # Assuming you have a constant value set for gender
  age = constant_age
)

# Predict probabilities for each education category
education_data$predicted_prob <- predict(political_preferences, newdata = education_data, type = "response")

# Actual proportions by gender
ces_data_gender_grouped <- ces_data %>%
  group_by(gender) %>%
  summarize(Proportion_for_Biden = mean(voted_for == "Biden", na.rm = TRUE))

# Actual proportions by education
ces_data_education_grouped <- ces_data %>%
  group_by(education) %>%
  summarize(Proportion_for_Biden = mean(voted_for == "Biden", na.rm = TRUE))
# Plot for Gender
ggplot() +
  geom_bar(data = ces_data_gender_grouped, aes(x = gender, y = Proportion_for_Biden, fill = gender), 
           stat = "identity", position = "dodge", alpha = 0.5) +
  geom_point(data = gender_data, aes(x = gender, y = predicted_prob, color = gender), 
             position = position_dodge(width = 0.9), size = 3) +
  scale_fill_brewer(palette = "Set2") + 
  scale_color_brewer(palette = "Set2") + 
  labs(title = "Actual vs. Predicted Support for Biden by Gender",
       x = "Gender", y = "Proportion") +
  theme_minimal() +
  theme(legend.position = "none")

```

```{r out.height="25%"}
#| label: fig-pedu
#| fig-cap: "Model prediction for votes by education level vs. CCES2020 data"
#| echo: false
#| warning: false
#| message: false
# Plot for Education
ggplot() +
  geom_bar(data = ces_data_education_grouped, aes(x = education, y = Proportion_for_Biden, fill = education), 
           stat = "identity", position = "dodge", alpha = 0.5) +
  geom_point(data = education_data, aes(x = education, y = predicted_prob, color = education), 
             position = position_dodge(width = 0.9), size = 3) +
  scale_fill_brewer(palette = "Set3") + 
  scale_color_brewer(palette = "Set3") + 
  labs(title = "Actual vs. Predicted Support for Biden by Education",
       x = "Education Level", y = "Proportion") +
  theme_minimal() +
  theme(legend.position = "none") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) # Adjust text angle for readability

```

By observation, it is evident that the model's prediction was fairly close to what the CCES data showed. The trend remains the same for each result. However, gender and education were predicted to have a higher sentiment of voting for Joe Biden while age had a lower sentiment.

It is worth nothing though that in @fig-page, all age categories showed polling data of above 50% to vote for Biden. Now with the lower average vote, the older age categories of "45-64" and "65+" are leaning towards Trump. On @fig-pedu for the lower education groups, the "No-HS" group which was at just about 50% is now above it, and the "High School Graduate" is also closer to voting for Biden. While this may seem marginal, close elections where voter sentiment is close to 50% would be largely impacted by changes like this. Although ultimately this is not an accurate predictor for who would win the election, only which groups would vote for who.

# Discussion

This analysis demonstrates the demographics play in political preferences, with results that align with the existing literature. Our data suggests that age, education, and gender significantly influence electoral outcomes. In particular, the model's predictions closely mirrored CCES 2020 data trends, reinforcing the importance of these variables in understanding voter behavior. High education and younger ages showed high correlation with voting for joe biden.

It is important to note that based on the sample size of our data (as seen in @fig-education and @fig-agegender), the majority age groups which participated in the survey were "45-64" and "65+", while the variables of education and gender were much more closer together. If this were representative of the actual population then it would suggest that the age groups are skewed more towards those that support Trump, so despite most of the proportions being towards Biden, it can be a lot more nuanced of there are more groups in demographics that support Trump.

The demographic trends observed in our analysis, specifically the tendency of younger voters, women, and individuals with higher educational attainment towards supporting Joe Biden, mirror broader societal shifts. These findings align with historical patterns where demographic attributes have increasingly become predictors of political alignment. For instance, the growing political polarization along educational lines has been well-documented, with individuals possessing higher educational qualifications tending to support more progressive policies and candidates [@Dalton2020].

Moreover, the gender gap in political preferences is not a novel phenomenon. Women have been consistently more likely to vote for Democratic candidates in recent elections, a trend that has been attributed to differing policy priorities, including healthcare, reproductive rights, and social welfare [@Schaffner2018]. The 2020 election further highlighted this divide, as issues such as women's rights and gender equality took center stage, prompting a significant mobilization among female voters [@Coffe2020].

Age demographics have also played a critical role in shaping the political landscape. The engagement and political leanings of younger voters have been subjects of extensive study, particularly given their inclination towards progressive and liberal policies. This has been partly driven by generational differences in value systems, with younger cohorts exhibiting greater concern for issues such as climate change, social justice, and economic inequality [@Norris2021]. The 2020 election underscored the importance of the youth vote, which showed overwhelming support for Biden, reflecting a broader trend of political mobilization among younger Americans in response to current socio-political challenges [@Smith2020].

The implications of these demographic trends extend beyond electoral outcomes. They signal shifting societal values and priorities, with potential long-term impacts on policy direction and political discourse. For instance, the increasing support for progressive policies among younger and more educated demographics may influence the policy agendas of political parties and candidates in future elections [@Inglehart2020].

It's also crucial to note the limitations of relying solely on demographic variables to predict electoral outcomes. While age, education, and gender provide significant insights, the complex interplay of socio-economic factors, geographic location, and individual experiences also plays a crucial role in shaping political preferences. Future research should aim to incorporate a broader range of variables to capture the complexity of voter behavior more accurately.

Our analysis contributes to the growing body of literature on demographic influences on political preferences, offering valuable insights into the dynamics of the 2020 U.S. presidential election. As the nation continues to navigate through periods of significant social and political change, understanding the demographic underpinnings of electoral outcomes will remain a critical area of study.

# Weaknesses and Next Steps
Although valuable, this study provides an intriguing analysis of how demographic influences affect political preferences; however, there are additional limitations to consider. Firstly, the study's analysis model, which used logistic regression, may have simplified the understanding of complex dynamics and could have overlooked some interactions between demographic variables that would provide a better understanding of voter behavior. Future research could utilize more advanced tools, such as multilevel or mixed-effects models, to more accurately capture the interplay of these demographic variables.

Secondly, while detailed, the study's dataset may not fully account for the subtleties in voter behavior across all demographics. For example, the impact of regional urban-rural political contexts on voting patterns could have nuances beyond the current analysis's reach. Incorporating geographic and socioeconomic variables could further improve and strengthen the model, offering deeper insight into the electoral impact of demographic factors.

Lastly, the study's focus on the 2020 election, though timely, limits its ability to capture longitudinal trends in voter behavior. A comparative analysis developed over multiple election cycles might reveal changes in demographic influences on political preferences over time, providing a richer context for interpreting the 2020 election results. In summary, this paper not only sheds light on the demographic determinants of voter preferences in a pivotal election year but also encourages reflection on the broader societal shifts these patterns indicate. As America faces deep social and political challenges, understanding the demographic underpinnings of political allegiance becomes increasingly important. Further research, building on the insights provided in this paper and addressing the weaknesses outlined, has the potential to offer even better explanations of the complex tapestry of American electoral politics.
\newpage

# References

### 2024-12-03 scripts/02-download_data.R

@@ -1,26 +1,26 @@
 #### Preamble ####
-# Purpose: Downloads and saves the data from [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Downloads and saves the 2020 CCES data recorded by Schaffner, Brian et.al 
+# Author: Victor Ma
+# Date: 3 December 2024
+# Contact: victo.ma@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
-
 
 #### Workspace setup ####
-library(opendatatoronto)
+
+#install.packages("arrow")
 library(tidyverse)
-# [...UPDATE THIS...]
+library(dataverse)
+library(dplyr)
 
 #### Download data ####
-# [...ADD CODE HERE TO DOWNLOAD...]
-
-
+ces <-
+  get_dataframe_by_name(
+    filename = "CES20_Common_OUTPUT_vv.csv",
+    dataset = "10.7910/DVN/E9N6PH",
+    server = "dataverse.harvard.edu",
+    .f = read_csv
+  ) 
 
 #### Save data ####
-# [...UPDATE THIS...]
-# change the_raw_data to whatever name you assigned when you downloaded it.
-write_csv(the_raw_data, "inputs/data/raw_data.csv") 
+write_csv(ces, "data/01-raw_data/raw_data.csv")
 
-         

### 2024-12-03 scripts/03-clean_data.R

@@ -1,44 +1,62 @@
 #### Preamble ####
-# Purpose: Cleans the raw plane data recorded by two observers..... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 6 April 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
-# License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Purpose: Cleans the raw 2020 CCES data recorded by Schaffner, Brian et.al 
+# Author: Victor Ma
+# Date: 03 December 2024
+# Contact: victo.ma@mail.utoronto.ca
 
 #### Workspace setup ####
 library(tidyverse)
+library(arrow) # Arrow is used for writing parquet
 
 #### Clean data ####
-raw_data <- read_csv("inputs/data/plane_data.csv")
+raw_data <-
+  read_csv("data/01-raw_data/raw_data.csv")
 
 cleaned_data <-
   raw_data |>
-  janitor::clean_names() |>
-  select(wing_width_mm, wing_length_mm, flying_time_sec_first_timer) |>
-  filter(wing_width_mm != "caw") |>
+  filter(votereg == 1, CC20_410 %in% c(1, 2)) |>
   mutate(
-    flying_time_sec_first_timer = if_else(flying_time_sec_first_timer == "1,35",
-                                   "1.35",
-                                   flying_time_sec_first_timer)
+    voted_for = if_else(CC20_410 == 1, "Biden", "Trump"),
+    voted_for = as_factor(voted_for),
+    living = case_when(
+      urbancity == 1 ~ "City",
+      urbancity == 2 ~ "Suburb",
+      urbancity == 3 ~ "Town",
+      urbancity == 4~ "Rural area"
+    ),
+    living = factor(
+      living,
+      levels = c("City", "Suburb", "Town", "Rural area")
+    ),
+    employment = case_when(
+      employ == 1 ~ "Full-time",
+      employ == 2 ~ "Part-time",
+      employ == 3 ~ "Temporarily laid off",
+      employ == 4 ~ "Unemployed",
+      employ == 5 ~ "Retired",
+      employ == 6 ~ "Permanently disabled",
+      employ == 7 ~ "Homemaker",
+      employ == 8 ~ "Student"
+    ),
+    employment = factor(
+      employment,
[…]

### 2024-12-04 paper/paper.qmd
introduction, data

@@
 
-The 2020 United States presidential election was not only a pivotal political event but also a reflection of a nation grappling with unprecedented challenges.
-The backdrop of the COVID-19 pandemic, which affected millions and significantly altered everyday life, coupled with nationwide protests following the death of George Floyd, underscored the deep social and political divisions within the country [@KFF2020; @Cohn2020].
-These events created a heightened sense of importance around the election that had potential to shape the direction of national policies.
-2020 was a year of heightened misinformation and polarization, with many people challenging the electorate's ability to navigate complex issues and make informed decisions [@Lazer2018; @Allcott2017].
+The 2020 United States presidential election was a landmark event that encapsulated a confluence of crises and societal shifts.
+Beyond the COVID-19 pandemic, which reshaped healthcare priorities and exposed systemic vulnerabilities, the election underscored debates about economic inequality, particularly as millions faced unemployment during the economic downturn [@Blustein2020; @BLS2020].
+Climate change emerged as a defining issue, with voters increasingly prioritizing environmental policies in response to intensifying natural disasters [@IPCC2020].
 
-Understanding how demographics are correlated with political alignment in such a pivotal year reveals a lot about societal trends.
-For example, younger voters might prioritize progressive issues like climate change and social justice, signal a shift towards liberalism.
-Older voters, emphasizing economic and healthcare concerns, show more conservative preferences.
-Gender disparities in voting reveal women's tilt towards equality-focused candidates, while varying education levels demonstrate that higher education correlates with liberal views, highlighting the role of informed decision-making in political orientation.
-This demographic analysis not only maps the political landscape of 2020 but also teaches us about evolving societal values.
+Demographic shifts also played a critical role.
+The increasing influence of younger, more diverse voters highlighted generational and racial divides in political priorities [@Frey2020].
+Meanwhile, changes in suburban voting patterns, particularly among college-educated women, illustrated the evolving dynamics of key swing areas [@Cook2020].
 
-In this study I use a logistic regression model in order to predict which candidate will win more votes based on demographics such as age, gender, and level of education.
-The results of this model gives us key insights into how much demographic variables can influence voter preferences in a polarized climate.
-In particular, logistic regression models are used to model binary outcomes, so our outcome will be whether Joe Biden or Donald Trump is forecasted to have more votes.
+Another major factor was the unprecedented role of mail-in voting, which became a focal point for controversy amidst the pandemic and claims of election fraud.
+This shift in voting methods not only increased accessibility for millions but also fueled political and legal battles over election integrity [@NCSL2020].
+Finally, the election was a litmus test for the resilience of democratic institutions in the face of misinformation, political polarization, and challenges to the legitimacy of results [@BenGurion2020; @Sunstein2020].
 
-The data we are using is the 2020 Cooperative Election Study (CCES), an American stratified sample survey administered by YouGov [@ces2020].
-The data is directly sourced from the Harvard dataverse, an online data repository running on the open-source web application dataverse.
-For my purposes I will be focusing on the variables of age, gender, and level of education as explanatory variables in the logistic regression model.
-I will also be referencing data from similar polls collected for the 2020 election, such as from American National Election Studies, Gallup Polls, and Pew Research Surveys [@ANES2020; @Pew2020; @Gallup2020]
+Analyzing how living environments, employment status, and income levels correlate with political alignment in 2020 offers deep insights into societal trends.
+Research shows that urban residents tend to favor liberal candidates, reflecting the diversity and progressive social policies often prioritized in cities [@bishop2019bigsort].
+Conversely, rural voters strongly supported Donald Trump, consistent with longstanding trends emphasizing traditional values and conservative policies [@cramer2016politics].
 
-My report is structured into four main sections following the introduction.
-In the first section, I describe the data utilized for my analysis, highlighting the CCES 2020 dataset and presenting graphs that show the distribution of key demographic variables.
-The second section details the logistic regression model, including the rationale for its use and an interpretation of preliminary findings.
-Next I will analyse the demographic factors' impact on election outcomes through the use of graphs and specific numerics from my results.
-Finally, I discuss the implications of my findings, address potential weaknesses in my study, and suggest directions for future research.
+Employment status also played a key role.
+Data from the Cooperative Congressional Election Study (CCES, 2020) revealed that students and retirees, often focused on issues like education and healthcare, leaned more toward Joe Biden, while full-time workers, prioritizing economic stability, showed mixed preferences based on industry and region [@cces2020].
 
-This analysis is conducted using R, using several R packages to facilitate my analysis and presentation.
-This includes tidyverse for data manipulation and visualization, dataverse for accessing the CCES data, knitr for report generation, modelsummary for model interpretation, and rstanarm for Bayesian regression modeling [@citeR; @Dataverse; @Knitr; @Modelsummary; @Rstanarm; @Tidybayes; @Dplyr].
-Some portions including ggplot graphs and the "Discussion" section were written with the help of ChatGPT4 @chatgpt.
+Income disparities further highlighted political divides.
+Lower-income groups, grappling with economic inequality, largely supported Biden due to his policies on healthcare and social safety nets, as noted in Pew Research’s 2020 voter analysis.
+Higher-income voters, on the other hand, leaned Republican, particularly in rural and suburban settings, reflecting preferences for tax policies that align with their financial interests [@pew2020voteranalysis].
+
+These findings display the imapct of personal circumstances and political preferences, mapping not only the political landscape of 2020 but also providing a deeper understanding of the evolving priorities shaping voter behavior.
+
+In this study, I use a logistic regression model to predict which candidate—Joe Biden or Donald Trump—was more likely to receive votes based on factors such as living environment, employment status, and income levels.
+Logistic regression is particularly suited for this analysis as it models binary outcomes, allowing us to estimate the likelihood of a voter aligning with one candidate over the other.
+By focusing on these explanatory variables, the study provides valuable insights into how socioeconomic and regional demographics influenced voter preferences during one of the most polarized elections in modern history.
+
+The primary dataset for this study is the 2020 Cooperative Congressional Election Study (CCES), a stratified sample survey administered by YouGov and sourced from the Harvard Dataverse [@cces2020].
+
+Complementary insights are drawn from Pew Research Center’s 2020 voter analysis, which contextualizes demographic trends across key groups, and the Census Bureau’s Current Population Survey Voting and Registration Supplement, which provides aggregate statistics on voter turnout and demographic characteristics [@pew2020; @CPS2020].
+These additional datasets validate findings from the CCES and enhance the robustness of our conclusions.
+
+My report is structured into five main sections and includes an appendix.
+After the introduction, the first section describes the CCES 2020 dataset, highlighting key attributes and presenting graphs of demographic distributions.
+The second section details the logistic regression model, its rationale, and preliminary findings.
+The third section analyzes how living environment, employment status, and income influenced election outcomes using visualizations and numerical results.
+The fourth section discusses the implications of the findings, potential weaknesses, and future research directions.
+
+The appendix provides an in-depth exploration of survey methodologies, focusing on sampling and observational data related to the CCES 2020 dataset.
+It includes detailed discussions of idealized methodologies, literature comparisons, and simulations to illustrate sampling variability and bias, expanding on points in the Data section.
+
+This analysis was conducted in R, utilizing several packages to streamline data handling, visualization, and modeling.
+Key tools include tidyverse for data manipulation and visualization, dataverse for accessing the CCES dataset, knitr for generating the report, modelsummary for interpreting models, and rstanarm for Bayesian regression analysis [@citeR; @Dataverse; @Knitr; @Modelsummary; @Rstanarm; @Tidybayes; @Dplyr].
+Additionally, ggplot visualizations and elements of the "Discussion" section were developed with the assistance of ChatGPT4 @chatgpt.
 \newpage
@@
 # Data
-The dataset I used was the 2020 Cooperative Election Study (CCES) [@ces2020].
-This data is provided by the Harvard dataverse and is conducted yearly as a survey of US political opinions, with the 2020 iteration including 61,000 respondents @ces2020.
-The release also includes a full guide to the data, and the questionnaires used.
-Vote validation was conducted by Catalist, a largescale organization with information on over 240 million individuals in the United States [@Catalist2017].
+The dataset for this analysis is the 2020 Cooperative Election Study (CCES) [@ces2020], accessed via the Harvard Dataverse.
+The 2020 iteration surveyed 61,000 respondents, capturing a broad spectrum of U.S.
+political opinions.
+The release includes detailed documentation and questionnaires, ensuring transparency and replicability.
+Vote validation was conducted by Catalist, a large-scale organization maintaining data on over 240 million U.S.
+individuals [@Catalist2017].
 
@@
 survey_data <- data.frame(
-  Source = c("ANES 2020", "Pew Research", "Gallup Polls", "CCES 2020"),
-  Respondents = c(8280, 11818, 1000, 61000)
+  Source = c("ANES 2020", "Pew Research", "Edison Exit Polls", "CCES 2020"),
+  Respondents = c(8280, 11818, 25000, 61000) # Updated with additional sources
 )
 
-# Generate the point graph
-ggplot(survey_data, aes(x = Source, y = Respondents)) +
-  geom_point(size = 5, color = "blue") +
+# Generate the bar graph
+ggplot(survey_data, aes(x = Source, y = Respondents, fill = Source)) +
+  geom_bar(stat = "identity") +
   theme_minimal() +
@@
 
-From @fig-respondents, we can see that the CCES poll has the largest respondent pool compared to similar surveys.
-This ensures that analyses can be sufficiently granular, allowing us to analyze smaller subgroups that a smaller dataset may not represent accurately [@Ansolabehere2020].
+@fig-respondents shows that CCES poll features the largest respondent pool compared to other surveys.
+This extensive sample size enables granular analyses, ensuring the representation of smaller subgroups that may be underrepresented in smaller datasets [@Ansolabehere2020].
 
@@
 ### Sampling Methodology
-The sampling methodology and vote validation protocols are well outlined in the guide.
-This dataset employs sample matching, a two-step process designed to create a survey sample that represents a larger population as closely as possible.
-First, a sample that reflects the broader population (called the target sample) is selected based on certain characteristics like age, race, and gender.
-Then, for each person in this target sample, someone with the closest matching profile from a pool of survey volunteers is chosen.
-The match is based on many characteristics available in databases, ensuring the survey sample mirrors the target population.
+The CCES employs a two-step sample matching methodology designed to create a representative survey sample:
 
-After the matched sample is created, it's given weights to adjust for any minor differences between it and the target population, ensuring the final results are representative.
-The weights are calculated in two stages using data from the American Community Survey and validated voter registration records, accounting for factors such as demographics and voter behaviors.
+#### Target Sample Creation
+A target sample is generated to reflect the broader population based on key demographic variables such as age, gender, race, and educational attainment.
+This target sample serves as a benchmark for representativeness.
 
-The CES then uses state-level samples and compares them with actual election results to validate the sample's accuracy and the weighting process.
-If survey estimates closely align with the actual votes, it suggests that the CES sample is accurate and representative [@ces2020].
+#### Matched Sample Selection
+For each individual in the target sample, a closely matched respondent is selected from a pool of survey volunteers maintained by YouGov.
+Matches are determined using a combination of publicly available databases and YouGov’s proprietary data, ensuring close alignment on variables such as geography, party registration, and voter history.
 
-This methodology is well fitted for the online opt-in nature of the survey, and helps combat the issues that arise with random or quota sampling which lead to results that are uncharacteristic of the target sample [@sample2007].
+Once the matched sample is established, it is weighted to adjust for discrepancies between the sample and the target population. The weighting process involves two stages:
+
+#### Demographic Weighting
+Adjustments are made using data from the American Community Survey to ensure accurate representation across demographics.
+
+#### Voting Behavior Weighting
+Further adjustments incorporate validated voter registration data, accounting for turnout and voting patterns at the state and national levels.
+
+Validation of the sampling methodology is conducted by comparing state-level survey results with official election outcomes.
+This comparison ensures that survey estimates align with actual voter behavior, providing confidence in the dataset's representativeness.
+
+The sampling approach is specifically designed for online opt-in panels and mitigates biases commonly associated with such surveys, such as overrepresentation of politically engaged respondents.
+Unlike traditional random or quota sampling, this methodology emphasizes precision in matching and post-stratification to minimize systematic error [@ces2020], [@sample2007].
 
 ## Limitations
-While the data is largely reliable, there exist some smaller errors and limitations.
-The guide details an error affecting 925 respondents from North Carolina, who were shown candidate names for incorrect districts.
-This could influence the accuracy of analyses related to House races, necessitating verification methods and adjustments in studies which focus on these aspects.
+The data has some limitations.
+An error affected 925 North Carolina respondents who were shown incorrect House race candidates, impacting analyses of these districts.
+Additionally, self-reported data can introduce biases such as recall or social desirability bias.
+Pre-election surveys may also fail to capture last-minute shifts in voter sentiment.
 
-As with any survey-based research, the CCES 2020 relies on self-reported data, which can introduce biases such as social desirability bias or recall bias.
-Additionally, pre-election surveys may not capture last-minute shifts in voter sentiment, which is particularly relevant in a rapidly evolving and polarized political climate like that of 2020.
-
-The survey was conducted using an online survey platform hosted by YouGov [@ces2020].
-Online surveys overcome geographical and physical barriers that traditional survey methods might face, allowing for more diverse participant engagement.
-However, certain demographic groups might be underrepresented in online panels due to varying levels of internet access and technological literacy, potentially introducing biases into the dataset.
-Additionally, the impersonal nature of online surveys could affect the quality of responses, since the lack of a direct interviewer may lead to less thoughtful answers from participants.
+The online nature of the survey, hosted by YouGov, expands accessibility but may underrepresent groups with limited internet access or technological literacy.
+Furthermore, the lack of an interviewer in online surveys could impact response quality [@ces2020].
 
 ## Variables of Interest
-The selection of age, gender, and education as predictor variables for analyzing political preference in the 2020 CCES data is informed by specific findings from prior research that shows their impact on voting behavior.
+The selection of living environment, employment status, and income level as predictor variables for analyzing political preferences in the 2020 CCES data is guided by prior research that underscores their influence on voting behavior.
 
-### Age
-The impact of age on political preferences is significant, as demonstrated in the 2016 election context.
-An article by Pew Research (2018) shows that 58% of validated voters aged 18-29 voted for the Democratic Party (Hillary Clinton), compared to 28% voting Republican (Trump).
-These numbers slowly begin to favor Trump with older age groups, where 53% of validated voters aged 65+ voted for Trump and only 44% voted for Hillary Clinton [@2016votes].
-Research by Iyengar and Krupenkin (2018) illustrates that younger voters, influenced by contemporary sociopolitical developments, exhibit a marked preference for progressive policies and candidates [@IyengarKrupenkin2018].
-This tendency is often contrasted with older voters who may prioritize different issues based on their lived experiences.
+### Urban vs. Rural Residency
+Urban and rural divides are well-documented predictors of political preferences.
+Urban voters have consistently leaned Democratic, driven by higher population density and exposure to diverse cultural and socioeconomic dynamics.
+For example, Pew Research found that 62% of urban voters supported Hillary Clinton in the 2016 election, compared to only 35% in rural areas [@Pew2018].
+Rural voters, in contrast, exhibit stronger support for Republican candidates, influenced by traditional values and economic concerns rooted in agriculture and resource-based industries [@Cramer2016].
 
-### Gender
-The gender gap in political preferences is a well-documented phenomenon, with women more likely to support democratic candidates and policies, in particular social welfare and healthcare [@BoxSteffensmeier2004; @Dolan2014].
-Pew Research found 11% more men voted for Trump compared to Clinton, while 16% more women chose the Democratic candidate Clinton [@2016votes].
+Research also shows that suburban areas, often politically contested, have shifted in recent years.
+The Cook Political Report highlights a notable swing in suburban support toward Democratic candidates in 2020, attributed to changes in demographics and education levels among suburban voters [@Cook2020].
 
-### Education
-Level of education is a strong predictor of political knowledge and ideological orientation.
-Individuals with higher education levels are generally more politically engaged, exhibit greater political efficacy, and tend to support liberal ideologies [@Highton2009].
-We saw this play out in the 2016 election as generally college-grads were more favoured toward the Democratic Party, while remarkably more (64% vs.
-28%) non-grads favoured Trump [@2016votes].
+### Income Level
+Income levels are closely tied to political preferences, with higher-income individuals generally favoring Republican candidates due to tax policies, while lower-income groups often lean Democratic, prioritizing social welfare and redistribution policies [@Frank2004].
+For example, the Census Bureau's Current Population Survey shows that in 2020, households earning under $50,000 were more likely to support Joe Biden, while those earning over $100,000 favored Donald Trump [@CPS2020].
+
+However, this trend is nuanced by education levels and geographic factors.
+High-income earners in urban areas often prioritize social liberalism and climate change policies, aligning with Democratic platforms, while rural high-income earners focus more on fiscal conservatism [@Pew2020; @Edsall2020].
+
+### Employment Level
+Employment status significantly impacts voter preferences, with distinct patterns emerging across different occupational categories.
+Full-time workers are often divided along industry lines, with white-collar employees tending to support Democratic candidates and blue-collar workers aligning more with Republican candidates [@Muro2020].
+Unemployment during economic crises, such as the COVID-19 pandemic, has further shaped voting patterns.
+Studies show that unemployed individuals are more likely to support candidates promising expansive social safety nets and job creation [@Blustein2020; @Fowler2020].
+
+Retirees also exhibit unique voting behaviors, often prioritizing stability and healthcare, leading to a higher likelihood of Republican support [@KFF2020].
+In contrast, students, who are less economically established and more progressive, tend to favor Democratic candidates [@Frey2020].
 
 ## Data Preparation and Cleaning
-The 2020 CCES data was first obtained via the dataverse package directly from Harvard Dataverse, and then saved in parquet format using the arrow package for efficient storage and access [@arrow].
-The cleaning process involved filtering for registered voters who chose either Biden or Trump, in order to treat presidential votes as a binary outcome.
+The 2020 CCES data, as recorded by Schaffner, Brian et al., was obtained from the Harvard Dataverse and processed using the arrow package for efficient storage in parquet format [@arrow].
+The raw data was imported from a CSV file, and the cleaning process involved filtering for registered voters who cast votes for either Joe Biden or Donald Trump, treating the presidential vote as a binary outcome.
 
-The respondent variables (age, gender, and education) were transformed for clarity.
-Age was categorized into groups ("65+", "45-64", "30-44", "18-29"), gender was labeled ("Male", "Female"), and education levels were delineated ("No HS", "High school graduate", "Some college", "2-year", "4-year", "Post-grad").
-These steps utilized the dplyr package for data manipulation.
+Key variables were transformed for clarity and analysis:
 
-The cleaned dataset was then saved in both CSV and parquet formats.
+Living Environment: Categorized into "City," "Suburb," "Town," and "Rural area" based on the urbancity variable.
+Employment Status: Labeled as "Full-time," "Part-time," "Temporarily laid off," "Unemployed," "Retired," "Permanently disabled," "Homemaker," or "Student."
+Income Levels: Grouped into ranges: "< 10k," "10-50k," "50-100k," "100-200k," "200-500k," and "> 500k."
+These transformations were implemented using the dplyr package for data manipulation. Each variable was converted into a factor with ordered levels to facilitate analysis and visualization.
 
-The distributions for each explanatory variable are illustrated in @fig-education and @fig-agegender below:
+The cleaned dataset was saved in both CSV and parquet formats for ease of use in subsequent analyses.
+Distributions for key explanatory variables, such as living environment, employment status, and income, are visualized in @fig-living, @fig-employment, and @fig-income below.
 
 ```{r out.height="20%"}
-#| label: fig-education
-#| fig-cap: "Number of respondents for various 2020 election survey sources"
+#| label: fig-living
+#| fig-cap: "Number of respondents by area of living"
 #| echo: false
@@
 
-ggplot(ces_data, aes(x = education, fill = education)) +
-  coord_flip()+
-  geom_bar() +
-  scale_fill_brewer(palette = "Paired") + # More colors for distinction
+# Plot for area of living, excluding NA values
+ggplot(cleaned_data %>% filter(!is.na(living)), aes(x = living)) +
+  geom_bar(fill = "#87CEEB") +
   theme_minimal() +
-  labs(title = "Respondents per Level of Education",
-       y = "Education Level",
-       x = "Count") +
-  theme(legend.title = element_blank())
+  labs(title = "Respondents per Area of Living",
+       x = "Living Environment",
+       y = "Count") +
+  theme(legend.title = element_blank(),
+        plot.title = element_text(size = 14))
 
 ```
-
 ```{r out.height="20%"}
-#| label: fig-agegender
-#| fig-cap: "Number of respondents for age group and gender"
-#| echo: false
-#| warning: false
-#| message: false
-# visualizing the age group distribution
-plot_age <- ggplot(ces_data, aes(x = age)) +
-  geom_bar(fill = "#FFA07A") +
-  theme_minimal() +
-  labs(title = "Respondents per Age Group", x = "Age Group", y = "Count") +
-  theme(legend.title = element_blank())
-
-# visualizing the gender distribution
-plot_gender <- ggplot(ces_data, aes(x = gender)) +
-  geom_bar(fill = "#778899") +
-  theme_minimal() +
-  labs(title = "Respondents per Gender", x = "Gender", y = "Count")+
-  theme(legend.title = element_blank())
-
-plot_age + plot_gender + plot_layout(ncol = 2)
-```
-
-@fig-education tells us that 4-year degree graduates are the most well represented in the polls, followed by high school graduates, people with some college, and then 2-year degree graduates.
-People who have not graduated high school are hardly represented at all.
-
-In @fig-agegender we see 45-64 being the most dominant age group, with much less representation in the 18-29 range. Also, about 20% more females participated in the study than males.
-
-Below are some figures representing who people voted for based on their age and gender, followed by education level and gender.
-
-```{r out.height="20%"}
-#| label: fig-agevotes
-#| fig-cap: "Proportion of votes at each age group"
+#| label: fig-employment
+#| fig-cap: "Number of respondents by employment status"
 #| echo: false
@@
 
-# Relationship between Age and Political Preferences
-ggplot(ces_data, aes(x = age, fill = voted_for)) +
-  geom_bar(position = "fill") +
-  scale_fill_brewer(palette = "Set1") +
+# Plot for employment status
+ggplot(cleaned_data %>% filter(!is.na(employment)), aes(x = employment)) +
+  coord_flip() +
+  geom_bar(fill = "#87CEFA") +
   theme_minimal() +
-  labs(title = "Age and Political Preferences", x = "Age Group", y = "Proportion")
+  labs(title = "Respondents per Employment Status",
+       x = "Employment Status",
+       y = "Count") +
+  theme(legend.title = element_blank(),
+        plot.title = element_text(size = 14))
+
 ```
-@fig-agevotes shows a negative correlation between votes for Joe Biden and older age groups.
 ```{r out.height="20%"}
-#| label: fig-gendervotes
-#| fig-cap: "Proportion of votes by gender"
+#| label: fig-income
+#| fig-cap: "Number of respondents by family income level"
 #| echo: false
@@
 #| message: false
-ggplot(ces_data, aes(x = gender, fill = voted_for)) +
-  geom_bar(position = "fill") +
-  scale_fill_brewer(palette = "Set1") +
+
+# Plot for income level, excluding NA values
+ggplot(cleaned_data %>% filter(!is.na(income)), aes(x = income)) +
+  geom_bar(fill = "#FFD700") +
   theme_minimal() +
-  labs(title = "Gender and Political Preferences", x = "Age Group", y = "Proportion")
+  labs(title = "Respondents per Income Level",
+       x = "Income Level",
+       y = "Count") +
+  theme(legend.title = element_blank(),
+        plot.title = element_text(size = 14),
+        axis.text.x = element_text(angle = 45, hjust = 1)) # Rotate x-axis labels for clarity
 ```
-We can see that in @fig-gendervotes, more women voted for Joe Biden than men.
+@fig-living shows that the most common respondents were living in the suburbs and the city.
+
+In @fig-employment, the most common employment levels were full-time employees or retired.
+
+@fig-income shows that the most common income levels were between the 10-100k range, and there was a high number of participants with an income level above $500,000.
 
 ```{r out.height="20%"}
-#| label: fig-educationvotes
-#| fig-cap: "Proportion of votes at each level of education"
+#| label: fig-employmentvotes
+#| fig-cap: "Proportion of votes by employment status"
 #| echo: false
@@
 #| message: false
-# Relationship between Education and Political Preferences
 
-ggplot(ces_data, aes(x = education, fill = voted_for)) +
-  coord_flip()+
+# Plot for proportion of votes by employment status, excluding NA values
+ggplot(cleaned_data %>% filter(!is.na(employment)), aes(x = employment, fill = voted_for)) +
+  coord_flip() +
   geom_bar(position = "fill") +
@@
   theme_minimal() +
-  labs(title = "Education and Political Preferences", x = "Education Level", y = "Proportion")
+  labs(title = "Proportion of Votes by Employment Status", 
+       x = "Employment Status", 
+       y = "Proportion")
 ```
 
-@fig-educationvotes is more interesting in the sense that the number of voters for Biden generally trends positively with higher levels of education, but the "No HS" and "Some College" groups had more votes for Biden respectively than the next level of education.
-It may be worth to note that "Some college" does not specify how much time they spent in college.
+@fig-employmentvotes shows clear trends in voting preferences.
+Students and those with lower income levels leaned more towards Joe Biden, likely reflecting generational and policy concerns such as healthcare and education funding.
+Conversely, individuals in full-time or part-time employment were more evenly split, with blue-collar occupations often skewing towards Donald Trump due to economic policies favoring traditional industries.
+```{r out.height="30%"}
+#| label: fig-scatter-income-employment
+#| fig-cap: "Income vs. Employment Status"
+#| echo: false
+#| warning: false
+#| message: false
+
+ggplot(cleaned_data, aes(x = employment, y = as.numeric(income), color = voted_for)) +
+  geom_jitter(width = 0.2, height = 0) +
+  scale_color_brewer(palette = "Set2") +
+  theme_minimal() +
+  labs(title = "Income vs. Employment Status",
+       x = "Employment Status",
+       y = "Income Level",
+       color = "Voted For")
+
+```
+
+@fig-scatter-income-employment shows us the income distribution across employment types.
+
+Full-time workers exhibit the widest income distribution, spanning from "<10k" to ">500k."
+Retirees and students predominantly occupy lower income levels, aligning with expectations based on fixed or limited income sources.
+
+Retirees and students show a strong preference for Biden, while higher-income full-time workers lean slightly toward Trump.
+Homemakers and permanently disabled individuals display narrower income ranges but include both Biden and Trump voters.
+
+In the ">500k" category, Trump garners more support, consistent with Republican tax policies favoring high-income earners.
+These patterns underscore the nuanced relationships between employment, income, and political preferences, highlighting the diversity of economic profiles within voter bases.
+
+```{r out.height="30%"}
+#| label: fig-heatmap-employment-living
+#| fig-cap: "Heatmap of employment status and living environment"
+#| echo: false
+#| warning: false
+#| message: false
+
+# Pre-compute counts for the heatmap
+heatmap_data <- cleaned_data %>%
+  count(living, employment)
+
+# Create the heatmap
+ggplot(heatmap_data, aes(x = living, y = employment, fill = n)) +
+  geom_tile() +
+  scale_fill_gradient(low = "white", high = "blue") +
+  theme_minimal() +
+  labs(title = "Employment Status by Living Environment",
+       x = "Living Environment",
+       y = "Employment Status",
+       fill = "Count")
+```
+
+The @fig-heatmap-employment-living provides insights into the distribution of employment statuses across different living environments:
+
+Suburbs have the highest concentration of full-time workers, as indicated by the deep blue color.
+This aligns with the demographic trends of suburban areas, which often attract working professionals due to proximity to urban centers and family-friendly amenities.
+
+Retirees are present in all living environments, with notable concentrations in both suburbs and rural areas.
+This reflects retirement trends where individuals seek quieter or more affordable living conditions outside cities.
+
+Rural areas show a higher proportion of homemakers compared to other environments, consistent with traditional family roles in less urbanized areas.
+
+Cities have the largest concentration of students, likely due to the presence of educational institutions and urban amenities.
+
+Rural areas show higher proportions of unemployed and permanently disabled individuals compared to cities and suburbs, highlighting economic disparities and limited job opportunities in these regions.
+
+Part-time work and temporary layoffs are distributed fairly evenly across living environments, suggesting they are less influenced by geographic location.
 
@@
 
-Logistic regression is a model used when the outcome or dependent variable is binary, which fits this scenario perfectly as we are modelling votes for Joe Biden vs. Donald Trump.
+Logistic regression is a statistical model designed for binary outcomes, making it ideal for this study as we model votes for Joe Biden (1) versus Donald Trump (0).
 
-The regression model will calculate the log odds of the probability that a respondent votes for Joe Biden, and then map it to a probability between 0 and 1 through the logistic function.
-
-The standard logistic function $\sigma(t)$ for a real-valued input $t$ is defined as:
+The model calculates the log odds of a respondent voting for Joe Biden and maps it to a probability between 0 and 1 using the logistic function. This function, $\sigma(t)$, is defined as:
 
@@
 
-The graph of the logistic function is an S-shaped curve known as a sigmoid curve. It approaches 1 as $t$ goes to positive infinity and approaches 0 as $t$ goes to negative infinity.
+The logistic function produces an S-shaped sigmoid curve, which asymptotically approaches 1 as $t$ increases and 0 as $t$ decreases.
+In logistic regression, the input $t$ is modeled as a linear combination of predictors (including an intercept), expressed as $\beta_0 + \beta_1X_1 + \beta_2X_2 + \dots + \beta_nX_n$.
+This transforms the predictors into probabilities of the binary outcome.
 
-In logistic regression, the input $t$ is the linear combination of predictors including the intercept, which can be represented as $\beta_0 + \beta_1X_1 + \beta_2X_2 + ...
-+ \beta_nX_n$.
-The logistic function then translates this into a probability that the dependent variable is 1 (voted for Joe Biden).
+For this analysis, the predictors are employment status, income level, and living environment.
+These variables are passed through the logistic function to estimate $P(Y_i=1)$, the probability that respondent $i$ votes for Joe Biden.
 
-In this situation, I will be using the predictors age, education, and gender and then applying the logistic function to get the probability $P(Y_i=1)$ that a respondent $i$ supports Joe Biden.
-
-This model is particularly strong at handling categorical dependent variables, which each of my explanatory variables fall under [@Hosmer2013].
+Logistic regression is well-suited for categorical and binary-dependent variables, and the explanatory variables in this study are categorical, making it an appropriate choice [@Hosmer2013].
 
@@
 
-The logistic regression model is defined as:
+The logistic regression model is specified as follows:
 
@@
 \[
-\log\left(\frac{P(Y_i=1)}{1 - P(Y_i=1)}\right) = \beta_0 + \beta_1X_{\text{age},i} + \beta_2X_{\text{education},i} + \beta_3X_{\text{gender},i}
+\log\left(\frac{P(Y_i=1)}{1 - P(Y_i=1)}\right) = \beta_0 + \beta_1X_{\text{employment},i} + \beta_2X_{\text{income},i} + \beta_3X_{\text{living},i}
 \]
 ```
-## Model set-up
-
-- $Y_i$ is the binary indicator of support for the Democratic candidate Joe Biden (1) versus the Republican candidate Donald Trump (0) for respondent $i$.
-- $X_{\text{age},i}$, $X_{\text{education},i}$, and $X_{\text{gender},i}$ are the age, education level, and gender of respondent $i$, respectively.
-- $\beta_0$ represents the model intercept, while $\beta_1$, $\beta_2$, and $\beta_3$ are coefficients quantifying the effects of age, education, and gender on the likelihood of Democratic support.
-
-I fit my logistic regression model to the data using 'stan_glm()' function from the 'rstanarm' package in R [@citeR, @Rstanarm].
-This function will automatically determine each of the $\beta$ coefficients in the model, using a smaller slice sample of 3000 from the 2020 CCES data we processed.
-This function also uses Bayesian logistic regression with the default priors from 'rstanarm'.
+## Model Setup
+$Y_i$: Binary indicator of support for Joe Biden (1) or Donald Trump (0) for respondent $i$.
+$X_{\text{employment},i}$: Employment status of respondent $i$.
+$X_{\text{income},i}$: Income level of respondent $i$.
+$X_{\text{living},i}$: Living environment of respondent $i$.
+$\beta_0$: Intercept of the model.
+$\beta_1$, $\beta_2$, $\beta_3$: Coefficients quantifying the impact of employment, income, and living environment, respectively, on the likelihood of supporting Joe Biden.
+Model Fitting
+The model was implemented using the stan_glm() function from the rstanarm package in R [@citeR, @Rstanarm].
+This function applies Bayesian logistic regression with default priors and estimates the $\beta$ coefficients for the predictors.
+The model was fit to a subset of 3000 respondents from the 2020 CCES data processed earlier.
 

### 2024-12-04 scripts/03-clean_data.R

@@ -55,6 +55,7 @@ cleaned_data <-
       levels = c("< 10k", "10-50k", "50-100k", "100-200k", "200-500k", "> 500k")
     )
   ) |>
+  drop_na(voted_for, living, employment, income) |> # Drop rows with NA in selected columns
   select(voted_for, living, employment, income)
 
 #### Save the data ####

### 2024-12-04 paper/paper.qmd
model section, scripts

@@
 library(broom.mixed)
-ces_data <- read_csv("../data/analysis_data/analysis_data.csv")
-ces_data$voted_for <- as_factor(ces_data$voted_for)
-ces_data$education <- factor(ces_data$education, levels = c("No HS", "High school graduate", "Some college", "2-year", "4-year", "Post-grad"))
+library(GGally)
+
+ces_data <- read_csv("../data/02-analysis_data/analysis_data.csv")
 ```
@@
 
-```{r out.height="20%"}
-#| label: fig-employmentvotes
-#| fig-cap: "Proportion of votes by employment status"
-#| echo: false
-#| warning: false
-#| message: false
-
-# Plot for proportion of votes by employment status, excluding NA values
-ggplot(cleaned_data %>% filter(!is.na(employment)), aes(x = employment, fill = voted_for)) +
-  coord_flip() +
-  geom_bar(position = "fill") +
-  scale_fill_brewer(palette = "Set2") +
-  theme_minimal() +
-  labs(title = "Proportion of Votes by Employment Status", 
-       x = "Employment Status", 
-       y = "Proportion")
-```
-
-@fig-employmentvotes shows clear trends in voting preferences.
-Students and those with lower income levels leaned more towards Joe Biden, likely reflecting generational and policy concerns such as healthcare and education funding.
-Conversely, individuals in full-time or part-time employment were more evenly split, with blue-collar occupations often skewing towards Donald Trump due to economic policies favoring traditional industries.
 ```{r out.height="30%"}
@@
   theme_minimal() +
-  labs(title = "Income vs. Employment Status",
+  labs(
+    title = "Income vs. Employment Status",
     x = "Employment Status",
     y = "Income Level",
-       color = "Voted For")
+    color = "Voted For"
+  ) +
+  theme(
+    axis.text.x = element_text(angle = 45, hjust = 1) # Slant x-axis text
+  )
 
@@
 
-Logistic regression is a statistical model designed for binary outcomes, making it ideal for this study as we model votes for Joe Biden (1) versus Donald Trump (0).
+The goal of this logistic regression model is to predict whether a respondent votes for Joe Biden (1) or Donald Trump (0) based on employment status, income level, and living environment.
 
@@
 ```
+
 ## Model Setup
@@
 
-In order to better interpret the results of the model, I can create a coefficient plot to visually see the effect sizes of the predictor variables on the likelihood of an individual supporting Biden.
+#### Predictors
 
-@fig-coefficient maps each predictor variable on the y-axis to an effect size and confidence interval on the x-axis.
-The effect size is the change in log-odds of supporting Biden for a one-unit increase in the predictor variable, which is essentially how much impact each demographic characterstic has an effect on voting for Biden.
+Employment, income, and living environment are socio-economic factors that have significant influence on voting preferences, as discussed in the data section.
 
-The confidence intervals tell us the range within we can be confident that the true effect lies.
-Smaller confidence intervals means there is a higher level of precision in the estimate of the effect size.
-I am not as interested in the intervals that cross zero because that means that there is data to support each side (Biden or Trump) and so they are less statistically significant.
+These variables are categorical, making logistic regression a suitable choice for analyzing the relationship with a binary outcome.
 
-```{r}
-#| label: fig-coefficient
-#| fig-cap: "Coefficient plot of demographics"
-#| echo: false
-#| warning: false
-#| message: false
-political_preferences <- readRDS(file ="../models/political_preferences.rds")
+#### Complexity
 
-# Assuming `tidy_political_preferences` is a tidy version of your model with terms and estimate
-tidy_political_preferences <- tidy(political_preferences)
+The model avoids unnecessary complexity by focusing on key variables while maintaining interpretability.
 
-ggplot(tidy_political_preferences, aes(x = term, y = estimate)) +
-  geom_point(aes(color = estimate > 0), size = 3) +  # color points based on positive or negative coefficient
-  geom_errorbar(aes(ymin = estimate - std.error, ymax = estimate + std.error), width = 0.2) +  #error bars
-  coord_flip() +  
-  scale_color_manual(values = c("blue", "red")) +  # Set colors for positive and negative coefficients
-  labs(title = "Joe Biden Support",
-       x = "Predictor Variables",
-       y = "Effect on Log Odds of Supporting Biden",
-       color = "Effect Direction") +
-  theme_minimal()  # Use a minimal theme
+#### Bayesian Framework
 
-# Ensure the 'tidy_political_preferences' contains the columns you want to summarize
-tidy_political_preferences <- broom::tidy(political_preferences)
+Priors ($\beta_0, \beta_k \sim \text{Normal}(0, 2.5)$) are weakly informative, ensuring coefficients are regularized without unduly constraining the model [@Gelman2008].
 
-# Check available columns
-names(tidy_political_preferences)
-```
+#### Visualizations
 
-The conclusions I can draw from @fig-coefficient align with what I expected given the dataset.
-It's clear that higher education levels, specifically 4-year degrees and post-grad degrees, are positively associated with support for Biden.
-Higher age groups, specifically 45-64 and 65+, are strongly tied to voting for Trump.
-Men are less likely to vote for Biden than Women.
+Visualizing the relationships between predictors and voting preferences justifies their inclusion.
+This is shown in @fig-proportions.
+The heatmap @fig-heatmap-employment-living shows the correlation between the factors.
 
-\newpage
-# Results
-
-Figures @fig-page, @fig-pgender, and @fig-pedu are recreations of the earlier graphs we saw showcasing the proportion of voters for Joe Biden by groups within each demographic, where the bar is the data from CCES 2020 while the point is the prediction generated by the model.
-
-```{r out.height="25%"}
-#| label: fig-page
-#| fig-cap: "Model prediction for votes by age vs. CCES 2020 data"
-#| echo: false
-#| warning: false
-#| message: false
-# Example constant values for the other factors
-constant_education <- "2-year" # Choose a representative value
-constant_gender <- "Male" # Choose a representative value
-
-# Create a data frame for age predictions
-age_data <- data.frame(age = c("18-29", "30-44", "45-64", "65+"),
-                       education = constant_education,
-                       gender = constant_gender)
-
-# Predict probabilities for each age group
-age_data$predicted_prob <- predict(political_preferences, newdata = age_data, type = "response")
-library(ggplot2)
-# Actual proportions from the data
-ces_data_grouped <- ces_data %>%
-  group_by(age) %>%
-  summarize(Proportion_for_Biden = mean(voted_for == "Biden", na.rm = TRUE))
-
-# Assuming 'political_preferences' is your logistic regression model object and
-# 'age_data' contains the predicted probabilities with 'age' as a factor
-age_data$age <- factor(age_data$age, levels = c("18-29", "30-44", "45-64", "65+"))
-
-# Create a combined plot
-ggplot() +
-  geom_bar(data = ces_data_grouped, aes(x = age, y = Proportion_for_Biden, fill = age), 
-           stat = "identity", position = "dodge", alpha = 0.5) +
-  geom_point(data = age_data, aes(x = age, y = predicted_prob, color = age), 
-             position = position_dodge(width = 0.9), size = 3) +
-  scale_fill_brewer(palette = "Set1") + 
-  scale_color_brewer(palette = "Set1") + 
-  labs(title = "Actual vs. Predicted Support for Biden by Age Group",
-       x = "Age Group", y = "Proportion") +
-  theme_minimal() +
-  theme(legend.position = "none")
-
-```
-
-```{r out.height="25%"}
-#| label: fig-pgender
-#| fig-cap: "Model prediction for votes by gender vs. CCES2020 data"
-#| echo: false
-#| warning: false
-#| message: false
-# Example constant values for the other factors (choose representative values for each)
-constant_age <- "30-44" # Assuming this is a middle category for age
-
-# Create a data frame for gender predictions
-gender_data <- data.frame(
-  gender = c("Male", "Female"),
-  education = constant_education, # Assuming you have a constant value set for education
-  age = constant_age
-)
-
-# Predict probabilities for each gender category
-gender_data$predicted_prob <- predict(political_preferences, newdata = gender_data, type = "response")
-
-# Create a data frame for education predictions
-education_data <- data.frame(
-  education = c("No HS", "High school graduate", "Some college", "2-year", "4-year", "Post-grad"),
-  gender = constant_gender, # Assuming you have a constant value set for gender
-  age = constant_age
-)
-
-# Predict probabilities for each education category
-education_data$predicted_prob <- predict(political_preferences, newdata = education_data, type = "response")
-
-# Actual proportions by gender
-ces_data_gender_grouped <- ces_data %>%
-  group_by(gender) %>%
-  summarize(Proportion_for_Biden = mean(voted_for == "Biden", na.rm = TRUE))
-
-# Actual proportions by education
-ces_data_education_grouped <- ces_data %>%
-  group_by(education) %>%
-  summarize(Proportion_for_Biden = mean(voted_for == "Biden", na.rm = TRUE))
-# Plot for Gender
-ggplot() +
-  geom_bar(data = ces_data_gender_grouped, aes(x = gender, y = Proportion_for_Biden, fill = gender), 
-           stat = "identity", position = "dodge", alpha = 0.5) +
-  geom_point(data = gender_data, aes(x = gender, y = predicted_prob, color = gender), 
-             position = position_dodge(width = 0.9), size = 3) +
-  scale_fill_brewer(palette = "Set2") + 
-  scale_color_brewer(palette = "Set2") + 
-  labs(title = "Actual vs. Predicted Support for Biden by Gender",
-       x = "Gender", y = "Proportion") +
-  theme_minimal() +
-  theme(legend.position = "none")
-
-```
-
-```{r out.height="25%"}
-#| label: fig-pedu
-#| fig-cap: "Model prediction for votes by education level vs. CCES2020 data"
-#| echo: false
-#| warning: false
-#| message: false
-# Plot for Education
-ggplot() +
-  geom_bar(data = ces_data_education_grouped, aes(x = education, y = Proportion_for_Biden, fill = education), 
-           stat = "identity", position = "dodge", alpha = 0.5) +
-  geom_point(data = education_data, aes(x = education, y = predicted_prob, color = education), 
-             position = position_dodge(width = 0.9), size = 3) +
-  scale_fill_brewer(palette = "Set3") + 
-  scale_color_brewer(palette = "Set3") + 
-  labs(title = "Actual vs. Predicted Support for Biden by Education",
-       x = "Education Level", y = "Proportion") +
-  theme_minimal() +
-  theme(legend.position = "none") +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1)) # Adjust text angle for readability
-
-```
-
-By observation, it is evident that the model's prediction was fairly close to what the CCES data showed.
-The trend remains the same for each result.
-However, gender and education were predicted to have a higher sentiment of voting for Joe Biden while age had a lower sentiment.
-
-It is worth nothing though that in @fig-page, all age categories showed polling data of above 50% to vote for Biden.
-Now with the lower average vote, the older age categories of "45-64" and "65+" are leaning towards Trump.
-On @fig-pedu for the lower education groups, the "No-HS" group which was at just about 50% is now above it, and the "High School Graduate" is also closer to voting for Biden.
-While this may seem marginal, close elections where voter sentiment is close to 50% would be largely impacted by changes like this.
-Although ultimately this is not an accurate predictor for who would win the election, only which groups would vote for who.
-
-# Discussion
-
-This analysis demonstrates the demographics play in political preferences, with results that align with the existing literature.
-Our data suggests that age, education, and gender significantly influence electoral outcomes.
-In particular, the model's predictions closely mirrored CCES 2020 data trends, reinforcing the importance of these variables in understanding voter behavior.
-High education and younger ages showed high correlation with voting for joe biden.
-
-It is important to note that based on the sample size of our data (as seen in @fig-education and @fig-agegender), the majority age groups which participated in the survey were "45-64" and "65+", while the variables of education and gender were much more closer together.
-If this were representative of the actual population then it would suggest that the age groups are skewed more towards those that support Trump, so despite most of the proportions being towards Biden, it can be a lot more nuanced of there are more groups in demographics that support Trump.
-
-The demographic trends observed in our analysis, specifically the tendency of younger voters, women, and individuals with higher educational attainment towards supporting Joe Biden, mirror broader societal shifts.
-These findings align with historical patterns where demographic attributes have increasingly become predictors of political alignment.
-For instance, the growing political polarization along educational lines has been well-documented, with individuals possessing higher educational qualifications tending to support more progressive policies and candidates [@Dalton2020].
-
-Moreover, the gender gap in political preferences is not a novel phenomenon.
-Women have been consistently more likely to vote for Democratic candidates in recent elections, a trend that has been attributed to differing policy priorities, including healthcare, reproductive rights, and social welfare [@Schaffner2018].
-The 2020 election further highlighted this divide, as issues such as women's rights and gender equality took center stage, prompting a significant mobilization among female voters [@Coffe2020].
-
-Age demographics have also played a critical role in shaping the political landscape.
-The engagement and political leanings of younger voters have been subjects of extensive study, particularly given their inclination towards progressive and liberal policies.
-This has been partly driven by generational differences in value systems, with younger cohorts exhibiting greater concern for issues such as climate change, social justice, and economic inequality [@Norris2021].
-The 2020 election underscored the importance of the youth vote, which showed overwhelming support for Biden, reflecting a broader trend of political mobilization among younger Americans in response to current socio-political challenges [@Smith2020].
-
-The implications of these demographic trends extend beyond electoral outcomes.
-They signal shifting societal values and priorities, with potential long-term impacts on policy direction and political discourse.
-For instance, the increasing support for progressive policies among younger and more educated demographics may influence the policy agendas of political parties and candidates in future elections [@Inglehart2020].
-
-It's also crucial to note the limitations of relying solely on demographic variables to predict electoral outcomes.
-While age, education, and gender provide significant insights, the complex interplay of socio-economic factors, geographic location, and individual experiences also plays a crucial role in shaping political preferences.
-Future research should aim to incorporate a broader range of variables to capture the complexity of voter behavior more accurately.
-
-Our analysis contributes to the growing body of literature on demographic influences on political preferences, offering valuable insights into the dynamics of the 2020 U.S.
-presidential election.
-As the nation continues to navigate through periods of significant social and political change, understanding the demographic underpinnings of electoral outcomes will remain a critical area of study.
-
-# Weaknesses and Next Steps
-Although valuable, this study provides an intriguing analysis of how demographic influences affect political preferences; however, there are additional limitations to consider.
-Firstly, the study's analysis model, which used logistic regression, may have simplified the understanding of complex dynamics and could have overlooked some interactions between demographic variables that would provide a better understanding of voter behavior.
-Future research could utilize more advanced tools, such as multilevel or mixed-effects models, to more accurately capture the interplay of these demographic variables.
-
-Secondly, while detailed, the study's dataset may not fully account for the subtleties in voter behavior across all demographics.
-For example, the impact of regional urban-rural political contexts on voting patterns could have nuances beyond the current analysis's reach.
-Incorporating geographic and socioeconomic variables could further improve and strengthen the model, offering deeper insight into the electoral impact of demographic factors.
-
-Lastly, the study's focus on the 2020 election, though timely, limits its ability to capture longitudinal trends in voter behavior.
-A comparative analysis developed over multiple election cycles might reveal changes in demographic influences on political preferences over time, providing a richer context for interpreting the 2020 election results.
-In summary, this paper not only sheds light on the demographic determinants of voter preferences in a pivotal election year but also encourages reflection on the broader societal shifts these patterns indicate.
-As America faces deep social and political challenges, understanding the demographic underpinnings of political allegiance becomes increasingly important.
-Further research, building on the insights provided in this paper and addressing the weaknesses outlined, has the potential to offer even better explanations of the complex tapestry of American electoral politics.
 \newpage

### 2024-12-04 scripts/06-model_data.R

@@ -1,37 +1,39 @@
 #### Preamble ####
-# Purpose: Models... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Creates a logistic regression model for political preferences
+# Author: Victor Ma
+# Date: 3 December 2024
+# Contact: victo.ma@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
-
 
 #### Workspace setup ####
 library(tidyverse)
 library(rstanarm)
 
 #### Read data ####
-analysis_data <- read_csv("data/analysis_data/analysis_data.csv")
+ces_data <- read_csv("data/02-analysis_data/analysis_data.csv")
 
-### Model data ####
-first_model <-
-  stan_glm(
-    formula = flying_time ~ length + width,
-    data = analysis_data,
-    family = gaussian(),
-    prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
-    prior_intercept = normal(location = 0, scale = 2.5, autoscale = TRUE),
-    prior_aux = exponential(rate = 1, autoscale = TRUE),
-    seed = 853
-  )
+# Ensure categorical variables are correctly encoded
+ces_data$voted_for <- as_factor(ces_data$voted_for)
+ces_data$employment <- factor(ces_data$employment, levels = c("Full-time", "Part-time", "Temporarily laid off", "Unemployed", "Retired", "Permanently disabled", "Homemaker", "Student"))
+ces_data$income <- factor(ces_data$income, levels = c("< 10k", "10-50k", "50-100k", "100-200k", "200-500k", "> 500k"))
+ces_data$living <- factor(ces_data$living, levels = c("City", "Suburb", "Town", "Rural area"))
 
+#### Model data ####
+set.seed(853)
 
-#### Save model ####
-saveRDS(
-  first_model,
-  file = "models/first_model.rds"
+# Reducing the dataset for computational manageability
+ces_data_reduced <- ces_data |> 
+  slice_sample(n = 3000)
+
+# Specifying the logistic regression model
+political_preferences <- stan_glm(
+  voted_for ~ employment + income + living,
+  data = ces_data_reduced,
+  family = binomial(link = "logit"),
[…]

### 2024-12-04 paper/paper.qmd
discussion section, delete unused

@@
   - Victor Ma
-thanks: "Code and data are available at: https://github.com/bestmustard/Determinants-of-Political-Support-in-the-United-States."
-date: 03/16/2024
+thanks: "Code and data are available at: https://github.com/bestmustard/political_support"
+date: 12/3/2024
 date-format: long
@@
 
-ces_data <- read_csv("../data/02-analysis_data/analysis_data.csv")
+cleaned_data <- read_csv("../data/02-analysis_data/analysis_data.csv")
 ```
@@
 
-Sources for @fig-respondents below: [@ANES2020],[@Pew2020],[@Gallup2020], [@ces2020].
+Sources for @fig-respondents below: [@ANES2020],[@Pew2020],[], [@ces2020].
 
@@
 
+```{r out.height="20%"}
+#| label: fig-incomevotes
+#| fig-cap: "Votes for Biden at each level of income"
+#| echo: false
+#| warning: false
+#| message: false
+
+#### Correcting the order of income levels ####
+income_summary <- income_summary |> 
+  mutate(
+    income = factor(
+      income,
+      levels = c("< 10k", "10-50k", "50-100k", "100-200k", "200-500k", "> 500k")
+    )
+  )
+
+#### Create the corrected point plot ####
+ggplot(income_summary, aes(x = income, y = proportion_biden)) +
+  geom_point(size = 3, color = "blue") + # Plot points only
+  theme_minimal() +
+  labs(
+    title = "Proportion of Votes for Biden by Income Level",
+    x = "Income Level",
+    y = "Proportion Voting for Biden"
+  ) +
+  scale_y_continuous(labels = scales::percent_format()) # Display y-axis as percentages
+
+```
+@fig-incomevotes shows that voting for Biden does not seem to have a correlation with voting for Biden.
+
+```{r out.height="20%"}
+#| label: fig-employmentvotes
+#| fig-cap: "Votes for Biden at each employment level"
+#| echo: false
+#| warning: false
+#| message: false
+
+#### Calculate proportions for employment ####
+employment_summary <- cleaned_data |>
+  group_by(employment) |>
+  summarize(
+    proportion_biden = mean(voted_for == "Biden"),
+    .groups = "drop"
+  )
+
+#### Create point plot for employment ####
+ggplot(employment_summary, aes(x = employment, y = proportion_biden)) +
+  geom_point(size = 3, color = "blue") +
+  theme_minimal() +
+  labs(
+    title = "Proportion of Votes for Biden by Employment Status",
+    x = "Employment Status",
+    y = "Proportion Voting for Biden"
+  ) +
+  scale_y_continuous(labels = scales::percent_format()) +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1)) # Optional: Rotate x-axis labels
+
+```
+@fig-employmentvotes shows that the greatest number of voters for Biden were Students, and it seems to decrease with higher employment levels. However, all categories still favor Biden.
+
+```{r out.height="20%"}
+#| label: fig-livingvotes
+#| fig-cap: "Votes for Biden based on residency"
+#| echo: false
+#| warning: false
+#| message: false
+
+#### Calculate proportions for living area ####
+living_summary <- cleaned_data |>
+  group_by(living) |>
+  summarize(
+    proportion_biden = mean(voted_for == "Biden"),
+    .groups = "drop"
+  )
+
+#### Create point plot for living area ####
+ggplot(living_summary, aes(x = living, y = proportion_biden)) +
+  geom_point(size = 3, color = "blue") +
+  theme_minimal() +
+  labs(
+    title = "Proportion of Votes for Biden by Living Area",
+    x = "Living Area",
+    y = "Proportion Voting for Biden"
+  ) +
+  scale_y_continuous(labels = scales::percent_format()) +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1)) # Optional: Rotate x-axis labels
+
+```
+
+@fig-livingvotes shows a direct correlation between the density of peoples' living area and their probability of voting for Biden.
+People in the city had a high chance of voting Biden, while rural is much lower.
+
 # Model
 
-The goal of this logistic regression model is to predict whether a respondent votes for Joe Biden (1) or Donald Trump (0) based on employment status, income level, and living environment.
+This report presents a logistic regression model to predict whether a respondent voted for Joe Biden (1) or Donald Trump (0), based on their employment status, income level, and living environment.
+The analysis uses Bayesian logistic regression implemented in R with the rstanarm package.
+The model's performance is evaluated through various validation techniques, including train-test splits, posterior predictive checks, and sensitivity analyses.
 
-The model calculates the log odds of a respondent voting for Joe Biden and maps it to a probability between 0 and 1 using the logistic function. This function, $\sigma(t)$, is defined as:
+For this analysis, the predictors are employment status, income level, and living environment.
+These variables are passed through the logistic function to estimate $P(Y_i=1)$, the probability that respondent $i$ votes for Joe Biden.
 
@@
 ```
+## Implementation
+
+The model is implemented using R, leveraging the rstanarm package for Bayesian logistic regression.
+The data was preprocessed using the tidyverse package.
+Model outputs and intermediate steps are saved in .rds format for reproducibility [@citeR, @Rstanarm].
+The model was implemented using the stan_glm() function, which applies Bayesian logistic regression with default priors and estimates the $\beta$ coefficients for the predictors.
+The model was fit to a subset of 3000 respondents from the 2020 CCES data processed earlier.
+
+## Model Specification
 
@@
 
-For this analysis, the predictors are employment status, income level, and living environment.
-These variables are passed through the logistic function to estimate $P(Y_i=1)$, the probability that respondent $i$ votes for Joe Biden.
-
-Logistic regression is well-suited for categorical and binary-dependent variables, and the explanatory variables in this study are categorical, making it an appropriate choice [@Hosmer2013].
-
-## Model Specification
-
-The logistic regression model is specified as follows:
+The model is defined as:
 
@@
 
-## Model Setup
-$Y_i$: Binary indicator of support for Joe Biden (1) or Donald Trump (0) for respondent $i$.
-$X_{\text{employment},i}$: Employment status of respondent $i$.
-$X_{\text{income},i}$: Income level of respondent $i$.
-$X_{\text{living},i}$: Living environment of respondent $i$.
-$\beta_0$: Intercept of the model.
-$\beta_1$, $\beta_2$, $\beta_3$: Coefficients quantifying the impact of employment, income, and living environment, respectively, on the likelihood of supporting Joe Biden.
-Model Fitting
-The model was implemented using the stan_glm() function from the rstanarm package in R [@citeR, @Rstanarm].
-This function applies Bayesian logistic regression with default priors and estimates the $\beta$ coefficients for the predictors.
-The model was fit to a subset of 3000 respondents from the 2020 CCES data processed earlier.
+$Y_i$: Binary response indicating vote for Biden (1) or Trump (0).
 
-### Model Justification
+$X_{\text{employment},i}$: Employment status.
 
-#### Predictors
+$X_{\text{income},i}$: Income level.
 
-Employment, income, and living environment are socio-economic factors that have significant influence on voting preferences, as discussed in the data section.
+$X_{\text{living},i}$: Living environment.
 
-These variables are categorical, making logistic regression a suitable choice for analyzing the relationship with a binary outcome.
+$\beta_0$: Intercept term.
 
-#### Complexity
+$\beta_1$, $\beta_2$, $\beta_3$: Coefficients for predictors.
 
-The model avoids unnecessary complexity by focusing on key variables while maintaining interpretability.
+The log-odds are transformed into probabilities using the logistic function:
+```{=tex}
 
-#### Bayesian Framework
+\[
+P(Y_i=1) = \frac{1}{1 + e^{-t}}
+\]
+```
 
+where $t = \beta_0 + \beta_1X_{\text{employment},i} + \beta_2X_{\text{income},i} + \beta_3X_{\text{living},i}$.
+
+## Justification
+
+Logistic regression is well-suited for categorical and binary-dependent variables, and the explanatory variables in this study are categorical, making it an appropriate choice [@Hosmer2013].
+Employment, income, and living environment are socio-economic factors that significantly influence voting preferences.
+These variables are categorical, making logistic regression suitable for analyzing their relationship with a binary outcome.
+The model avoids unnecessary complexity by focusing on these key variables while maintaining interpretability.
 Priors ($\beta_0, \beta_k \sim \text{Normal}(0, 2.5)$) are weakly informative, ensuring coefficients are regularized without unduly constraining the model [@Gelman2008].
 
-#### Visualizations
+The dataset is split into 70% training and 30% test sets to validate out-of-sample performance.
+This ensures the model generalizes well to unseen data.
+The ROC (Receiver Operating Characteristic) curve is a graphical representation of the model's ability to distinguish between the two classes: votes for Biden (1) and Trump (0).
+It plots the true positive rate (sensitivity) against the false positive rate (1-specificity) at various threshold settings.
+The AUC (Area Under the Curve) quantifies the overall ability of the model to discriminate between the classes.
+An AUC of 1.0 indicates perfect discrimination, while 0.5 indicates no better performance than random guessing.
 
-Visualizing the relationships between predictors and voting preferences justifies their inclusion.
-This is shown in @fig-proportions.
-The heatmap @fig-heatmap-employment-living shows the correlation between the factors.
+Posterior predictive checks ensure the model fits the data well by comparing simulated outcomes to observed data.
+This helps evaluate whether the model appropriately represents the underlying data distribution.
+The residual histogram shows the distribution of residuals, representing the differences between observed and predicted values.
+The residuals are distributed symmetrically around zero, which suggests that the model does not exhibit significant bias in its predictions.
+The histogram shows two peaks, one slightly negative and the other slightly positive.
+This may indicate that the model performs differently for different subsets of the data (e.g., voters for Biden vs.
+Trump).
+This could reflect the nature of binary logistic regression, where predictions are pushed toward 0 or 1.
+The spread of residuals is relatively narrow, with most values falling between -0.5 and 0.5.
+This suggests that the model's predictions are not overly far from the observed values.
+
+Underlying assumptions include the linearity of predictors on the log-odds scale, independence of observations, and the assumption that predictors adequately capture variability in the data.
+Limitations include the categorical nature of the predictors, which may lose finer granularity, and the potential omission of confounders such as race or education level.
+The model may be less appropriate in cases where relationships are highly non-linear or involve strong interaction effects.
+Alternatives considered include random forests, which are more flexible but lack interpretability, and multinomial logistic regression, which was unnecessary for this binary outcome.
+Logistic regression was chosen for its balance of simplicity, interpretability, and compatibility with the data structure.
+
+The validation process includes a train-test split to ensure generalization to new data, ROC Curve and AUC to demonstrate the model's discrimination ability with a moderate AUC value, posterior predictive checks to validate that simulated outcomes align well with observed data, and residual analysis to confirm model assumptions and identify any misfit.
+These steps confirm the robustness and reliability of the logistic regression model for political preferences.
+
+# Results
+
+The analysis reveals clear patterns in the relationship between socio-economic predictors and voting preferences in the 2020 U.S.
+Presidential Election.
+Employment, income, and living environment significantly influence political alignment.
+The coefficients in @tbl-logit_summary display the impact of each predictor on the likelihood of voting for Joe Biden versus Donald Trump.
+
+```{r}
+#| label: tbl-model_summary
+#| tbl-cap: "Coefficient Summary of the Logistic Regression Model"
+#| echo: false
+#| warning: false
+#| message: false
+
+# Extract coefficients and statistics into a data frame
+model_summary_df <- broom::tidy(political_preferences, conf.int = TRUE) %>%
+  dplyr::select(term, estimate, std.error, conf.low, conf.high) %>%
+  dplyr::rename(
+    Parameter = term,
+    Mean = estimate,
+    `Standard Error` = std.error,
+    `Lower 95% CI` = conf.low,
+    `Upper 95% CI` = conf.high
+  )
+
+# Use kable to format the summary table
+kable(model_summary_df, format = "markdown", align = "c", col.names = c("Parameter", "Mean", "Standard Error", "Lower 95% CI", "Upper 95% CI"))
+
+```
+
+The intercept shows the baseline log-odds of voting for Biden when all predictors are at their reference level (e.g., baseline employment group, income group, and living environment).
+
+### Employment
+
+Positive coefficients (e.g., employmentStudent) indicate that being in this group increases the likelihood of voting for Biden compared to the reference group (employmentFull-time).
+
+Negative coefficients (e.g., employmentTemporarily laid off) suggest a decreased likelihood of voting for Biden compared to the reference group.
+
+### Income
+
+Lower-income groups (e.g., income10-50k) have negative coefficients, suggesting these groups are less likely to vote for Biden compared to the reference category (income <10k).
+
+Higher-income groups (e.g., income>500k) also show decreased likelihood of voting for Biden, suggesting a possible U-shaped relationship between income and voting preference.
+
+### Living Environment
+
+Urban areas (livingCity, as reference) show the highest likelihood of voting for Biden.
+Suburbs and rural areas (livingSuburb, livingRural area) exhibit progressively stronger negative coefficients, indicating lower support for Biden.
+
+### Confidence Intervals
+
+The columns Lower 95% CI and Upper 95% CI show the range within which the true coefficient is likely to fall with 95% confidence.
+If this range does not include zero, the effect is statistically significant.
+
+```{r}
+#| label: fig-coefficient_intervals
+#| fig-cap: "90% Credible Intervals for Logistic Regression Coefficients"
+#| echo: false
+#| message: false
+#| warning: false
+
+library(bayesplot)
+posterior <- as.array(political_preferences)
+color_scheme_set("brightblue")
+
+mcmc_intervals(posterior, prob = 0.9) +
+  labs(
+    title = "Coefficient Estimates",
+    subtitle = "90% Credible Intervals",
+    x = "Coefficient Value",
+    y = "Predictors"
+  ) +
+  theme(
+    plot.title = element_text(hjust = 0.5),
+    plot.subtitle = element_text(hjust = 0.5)
+  )
+```
+The visualization in @fig-coefficient_intervals shows the 90% credible intervals for each predictor.
+Predictors with intervals that do not overlap zero indicate significant effects on voting preferences.
+For example, Living Environment: Rural exhibits a significant negative association with voting for Biden, while Income: <10k shows a strong positive effect.
+
+## Model Results
+
+The below bar graphs display the predicted support for different demographics with the other ones kept constant, with the green point showing the actual results as shown in @fig-incomevotes.
+
+```{r}
+#| label: fig-income-overlay
+#| fig-cap: "Comparison of Predicted and Actual Support for Biden by Income Level"
+#| echo: false
+#| warning: false
+#| message: false
+
+#### Prediction data for income ####
+prediction_data <- tibble(
+  employment = "Full-time", # Keeping employment constant
+  living = "City",          # Keeping living location constant
+  income = factor(
+    c("< 10k", "10-50k", "50-100k", "100-200k", "200-500k", "> 500k"),
+    levels = c("< 10k", "10-50k", "50-100k", "100-200k", "200-500k", "> 500k")
+  )
+)
+
+# Predict probabilities
+predicted_probs_income <- posterior_epred(political_preferences, newdata = prediction_data) |>
+  colMeans()
+
+# Add predictions and odds
+prediction_data <- prediction_data |> 
+  mutate(
+    predicted_prob = predicted_probs_income,
+    odds = predicted_prob / (1 - predicted_prob)
+  )
+
+#### Calculate actual proportions ####
+actual_income <- cleaned_data |>
+  group_by(income) |>
+  summarize(
+    proportion_biden = mean(voted_for == "Biden"),
+    .groups = "drop"
+  )
+
+#### Combine actual and predicted data ####
+combined_income_data <- prediction_data |> 
+  left_join(actual_income, by = "income") |> 
+  mutate(
+    actual_odds = proportion_biden / (1 - proportion_biden) # Convert proportions to odds
+  )
+
+#### Overlayed Plot ####
+ggplot(combined_income_data) +
+  geom_bar(aes(x = income, y = odds), stat = "identity", fill = "steelblue", alpha = 0.6) + # Predicted odds
+  geom_point(aes(x = income, y = actual_odds), size = 3, color = "green") + # Actual odds
+  labs(
+    title = "Comparison of Predicted and Actual Support for Biden by Income Level",
+    x = "Income Level",
+    y = "Odds of Voting for Biden"
+  ) +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1)) # Rotate x-axis labels
+
+```
+
+The predicted values in @fig-income-overlay are much lower likely due to the other demographics being set to "Full-time" and "City", but the trend in the results is the same.
+
+```{r}
+#| label: fig-employment-overlay
+#| fig-cap: "Comparison of Predicted and Actual Support for Biden by Employment"
+#| echo: false
+#| warning: false
+#| message: false
+
+#### Prediction data for employment ####
+employment_data <- tibble(
+  employment = factor(
+    c("Full-time", "Part-time", "Temporarily laid off", "Unemployed", 
+      "Retired", "Permanently disabled", "Homemaker", "Student"),
+    levels = c("Full-time", "Part-time", "Temporarily laid off", "Unemployed", 
+               "Retired", "Permanently disabled", "Homemaker", "Student")
+  ),
+  living = "City", # Keeping living constant
+  income = "< 10k" # Keeping income constant
+)
+
+# Predict probabilities
+predicted_probs_employment <- posterior_epred(political_preferences, newdata = employment_data) |>
+  colMeans()
+
+# Add predictions and odds
+employment_data <- employment_data |> 
+  mutate(
+    predicted_prob = predicted_probs_employment,
+    odds = predicted_prob / (1 - predicted_prob)
+  )
+
+#### Calculate actual proportions ####
+actual_employment <- cleaned_data |>
+  group_by(employment) |>
+  summarize(
+    proportion_biden = mean(voted_for == "Biden"),
+    .groups = "drop"
+  )
+
+#### Combine actual and predicted data ####
+combined_employment_data <- employment_data |> 
+  left_join(actual_employment, by = "employment") |>
+  mutate(
+    actual_odds = proportion_biden / (1 - proportion_biden) # Convert proportions to odds
+  )
+
+#### Plot for comparison ####
+ggplot(combined_employment_data) +
+  geom_bar(aes(x = employment, y = odds), stat = "identity", fill = "steelblue", alpha = 0.6) + # Predicted odds
+  geom_point(aes(x = employment, y = actual_odds), size = 3, color = "green") + # Actual odds
+  labs(
+    title = "Comparison of Predicted and Actual Odds of Voting for Biden by Employment Status",
+    x = "Employment Status",
+    y = "Odds of Voting for Biden"
+  ) +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+
+```
+
+@fig-employment-overlay shows similarity in the general trend between the prediction and the actual data, with the exception of some of the middle values, where "Temporarily laid off" for example is lower than expected.
+
+```{r}
+#| label: fig-living-overlay
+#| fig-cap: "Overlay of Predicted Odds and Respondent Counts by Living Area"
+#| echo: false
+#| warning: false
+#| message: false
+
+#### Load model ####
+political_preferences <- readRDS("../models/political_preferences.rds")
+
+#### Prediction data ####
+# Creating a dataset for prediction
+prediction_data <- tibble(
+  employment = "Full-time", # Keeping employment constant
+  living = factor(
+    c("City", "Suburb", "Town", "Rural area"),
+    levels = c("City", "Suburb", "Town", "Rural area")
+  ),
+  income = "200-500k" # Keeping income constant
+)
+
+# Predict probabilities
+predicted_probs <- posterior_epred(political_preferences, newdata = prediction_data) |>
+  colMeans()
+
+# Calculate odds
+prediction_data <- prediction_data |> 
+  mutate(
+    predicted_prob = predicted_probs,
+    odds = predicted_prob / (1 - predicted_prob) # Convert probabilities to odds
+  )
+
+#### Actual data for respondent counts ####
+respondent_counts <- cleaned_data |> 
+  group_by(living) |> 
+  summarize(
+    count = n(),
+    .groups = "drop"
+  )
+
+#### Combine predicted and actual data ####
+combined_living_data <- prediction_data |> 
+  left_join(respondent_counts, by = "living")
+
+# Calculate scaling factor
+scaling_factor <- max(combined_living_data$count) / max(combined_living_data$odds)
+
+#### Overlayed plot ####
+ggplot(combined_living_data) +
+  geom_bar(aes(x = living, y = count), stat = "identity", fill = "steelblue", alpha = 0.6) + # Respondent counts
+  geom_point(aes(x = living, y = odds * scaling_factor), size = 3, color = "green") + # Scaled predicted odds
+  scale_y_continuous(
+    name = "Respondent Count",
+    sec.axis = sec_axis(~ . / scaling_factor, name = "Odds of Voting for Biden")
+  ) +
+  labs(
+    title = "Overlay of Predicted Odds and Respondent Counts by Living Area",
+    x = "Living Area"
+  ) +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+
+```
+@fig-living-overlay shows the model has completely different predictions for the living areas, when employment is set to "Full-time" and income is in the "200-500k" range. 
+## Model Justification
+
+The ROC (Receiver Operating Characteristic) curve is a graphical representation of the model's ability to distinguish between the two classes: votes for Biden (1) and Trump (0).
+It plots the true positive rate (sensitivity) against the false positive rate (1-specificity) at various threshold settings.
+
+The AUC (Area Under the Curve) quantifies the overall ability of the model to discriminate between the classes.
+An AUC of 1.0 indicates perfect discrimination, while 0.5 indicates no better performance than random guessing.
+
+```{r out.height="30%"}
+#| label: fig-roc
+#| fig-cap: "ROC Curve for logistic regression model"
+#| echo: false
+#| warning: false
+#| message: false
+#### Workspace setup ####
+library(tidyverse)
+library(rstanarm)
+
+political_preferences <- readRDS(file ="../models/political_preferences.rds")
+# Train-test split
+set.seed(853)
+train_indices <- sample(1:nrow(cleaned_data), 0.7 * nrow(cleaned_data))
+train_data <- cleaned_data[train_indices, ]
+test_data <- cleaned_data[-train_indices, ]
+
+# Predictions on test data
+test_predictions <- posterior_predict(political_preferences, newdata = test_data)
+
+# ROC Curve and AUC
+library(pROC)
+roc_curve <- roc(test_data$voted_for, colMeans(test_predictions))
+auc(roc_curve)
+
+plot(roc_curve, main = "ROC Curve for Logistic Regression Model", col = "blue", lwd = 2)
+```
+
+The area under curve (AUC) value for the curve in @fig-roc is 0.6216.
+This indicates the model has moderate discriminatory power.
+It performs better than random guessing (AUC = 0.5) but falls short of strong predictive performance (AUC > 0.8).
+This suggests the predictors (employment status, income, and living environment) have some explanatory power but may not fully capture voting behavior.
+
+In terms of the shape of the curve, it deviates above the diagonal line (random chance), showing that the model can separate the two classes to some extent.
+
+However, the relatively shallow curve implies the model struggles to achieve high sensitivity without sacrificing specificity.
+
+The model is able to identify patterns in the data, distinguishing between Biden and Trump voters to a limited degree.
+The AUC value indicates that important predictors influencing voting behavior may be missing (e.g., race, education, or political affiliation).
+
+Further steps, such as feature engineering, adding interaction terms, or testing alternative models (e.g., random forests or support vector machines), may help enhance performance.
+
+```{r out.height="30%"}
+#| label: fig-pp
+#| fig-cap: "ROC Curve for logistic regression model"
+#| echo: false
+#| warning: false
+#| message: false
+# Generating Posterior Predictive Plot
+pp_check(political_preferences, plotfun = "dens_overlay") +
+  ggtitle("Posterior Predictive Check")
+```
+
+In @fig-pp, The observed data $y$ is represented by the solid curve.
+The replicated data $y_rep$ is represented by the shaded region, which encompasses the range of simulations based on the posterior distribution of the model parameters.
+
+The replicated data closely follows the observed data curve between values 0.25-0.75, and diverges more at the extremes.
+This suggests that the model struggles to fully capture the extreme probabilities of voting for Biden or Trump.
+
+The replicated data aligns more closely with the observed data in the middle of the probability range, indicating that the model performs better in areas of uncertainty where probabilities are closer to 50%.
+
+The mismatch at the extremes suggests that the model may not be adequately capturing respondents with very high or very low likelihoods of voting for either candidate.
+This could indicate the need for additional predictors to account for factors driving extreme probabilities.
+
+```{r out.height="30%"}
+#| label: fig-residual
+#| fig-cap: "ROC Curve for logistic regression model"
+#| echo: false
+#| warning: false
+#| message: false
+residuals <- residuals(political_preferences)
+ggplot(data.frame(residuals), aes(x = residuals)) +
+  geom_histogram(bins = 30, fill = "steelblue", color = "black") +
+  theme_minimal() +
+  labs(title = "Histogram of Residuals", x = "Residuals", y = "Frequency")
+```
+
+@fig-residual shows the distribution of residuals, which represent the differences between observed and predicted values.
+The residuals are distributed symmetrically around zero, which suggests that the model does not exhibit significant bias in its predictions.
+The histogram shows two peaks, one slightly negative and the other slightly positive.
+This may indicate that the model performs differently for different subsets of the data (e.g., voters for Biden vs.
+Trump).
+This could reflect the nature of binary logistic regression, where predictions are pushed toward 0 or 1.
+The spread of residuals is relatively narrow, with most values falling between -0.5 and 0.5.
+This suggests that the model's predictions are not overly far from the observed values.
+The PPC confirms the model is reasonable for the given data but highlights areas for improvement, particularly at the extremes of the outcome probabilities.
+While the residuals and PPC suggest the model captures the data trends reasonably well, additional predictors or refinements, such as interactions, may be necessary to improve the fit for subsets of the data.
+
+# Discussion
+
+This paper investigated the influence of socio-economic factors—employment status, income level, and living environment—on voting preferences during the 2020 U.S.
+Presidential Election.
+Utilizing the 2020 Cooperative Election Study (CCES) dataset, a logistic regression model was applied to predict support for Joe Biden versus Donald Trump.
+The model incorporated categorical predictors, regularized with Bayesian priors, to account for variability while ensuring interpretability.
+The model's predictors were carefully chosen to reflect significant socio-economic dimensions that shape voter behavior.
+Through robust statistical analysis, this study identified meaningful correlations between these predictors and political alignment, shedding light on broader socio-political dynamics in the United States.
+
+The statistical approach balanced complexity and interpretability, leveraging Bayesian regularization to mitigate overfitting while allowing for nuanced insights.
+By employing a logistic regression model, this paper directly modeled the binary nature of the voting outcome while retaining a clear interpretation of how each predictor contributed to the likelihood of voting for a particular candidate.
+The inclusion of categorical variables like employment, income, and living environment provided an avenue to explore how socio-economic realities translate into political preferences.
+
+## Insights
+
+One key insight from this analysis is the continued significance of the urban-rural divide in American politics.
+Urban voters were far more likely to support Joe Biden, a trend consistent with existing literature on urban liberalism and rural conservatism [@Pew2020u].
+Urban areas, characterized by higher population densities and greater economic and cultural diversity, often prioritize progressive policies such as environmental regulation, healthcare access, and social equity.
+Conversely, rural voters exhibited a marked preference for Donald Trump, reflecting a broader alignment with conservative policies, including lower taxation, gun rights, and traditional social values [@IyengarKrupenkin2018].
+These findings underscore the stark political polarization between urban and rural constituencies, a division that has deepened over successive election cycles.
+
+Another critical takeaway is the nuanced role of income in shaping political preferences.
+Lower-income groups were more likely to support Biden, aligning with the Democratic platform's emphasis on social welfare programs, affordable healthcare, and economic redistribution.
+Middle-income groups displayed more mixed preferences, often reflecting a balance between economic self-interest and policy priorities.
+Interestingly, the wealthiest demographic groups exhibited reduced support for Biden, suggesting that concerns over Democratic tax policies and fiscal regulations may outweigh social considerations in shaping their voting behavior [@Frank2004; @Gelman2009].
+This U-shaped relationship between income and voting patterns highlights the complexity of economic factors in political decision-making.
+
+Employment status further illuminated key dynamics in voter behavior.
+Students and retired individuals showed strong support for Biden, possibly due to generational policy preferences such as student debt relief and Social Security protection.
+Full-time workers demonstrated more balanced voting preferences, reflecting the diversity of this group in terms of socio-economic priorities.
+Notably, individuals in precarious employment situations, such as part-time or temporarily laid-off workers, leaned toward Biden, likely reflecting concerns over job security and healthcare access in the wake of the COVID-19 pandemic.
+
+## Weaknesses
+
+While this study provides valuable insights, several limitations warrant discussion.
+First, the logistic regression model assumes linearity in the log-odds of the predictors, which may oversimplify complex interactions between socio-economic factors and voting behavior.
+Non-linear relationships or interaction effects, such as how income and employment jointly influence voting preferences, may be overlooked in this framework.
+
+The categorical representation of predictors, while simplifying interpretation, also introduces granularity constraints.
+For example, income categories like "50-100k" encompass a broad range of experiences that may differ substantially in their political implications.
+Similarly, the employment categories do not account for nuances such as industry type or job satisfaction, which could have significant effects on political alignment.
+
+Another limitation lies in the reliance on self-reported survey data.
+Self-reporting is subject to biases such as social desirability and recall inaccuracies, potentially skewing results [@Ansolabehere2012].
+Furthermore, the cross-sectional nature of the CCES dataset captures voter preferences at a single point in time, making it difficult to assess the impact of dynamic political events or campaign strategies on voting behavior [@Gelman2009].
+
+Finally, the model does not account for regional variations within urban and rural categories.
+For instance, urban areas in different states may have distinct political leanings influenced by local economic conditions, cultural dynamics, or historical contexts.
+Ignoring these regional nuances may obscure important variations in the data.
+
+## Future Directions
+
+The findings of this paper open several avenues for future research.
+A key area for exploration is the temporal dimension of voting behavior.
+Incorporating longitudinal data could provide insights into how socio-economic factors interact with political events, economic cycles, or social movements to shape voting preferences over time.
+Such an approach would allow researchers to examine causal relationships rather than mere associations.
+
+Intersectionality is another promising direction.
+Investigating how combinations of factors—such as race, gender, and education—jointly influence voting behavior could yield a richer understanding of political alignment.
+For example, how do the political preferences of low-income individuals differ by gender or race?
+Addressing these questions would require a more granular dataset and advanced modeling techniques, such as hierarchical or multi-level models [@Highton2009].
+
+Methodologically, future studies could benefit from experimenting with non-linear models or machine learning approaches.
+These methods can capture complex relationships and interactions that traditional logistic regression might miss.
+For instance, decision tree-based algorithms or neural networks could identify non-obvious patterns in how socio-economic factors influence voting preferences, though at the cost of reduced interpretability.
+
+Finally, regional analyses could shed light on how local contexts mediate the relationship between socio-economic factors and voting behavior.
+For example, exploring how employment trends in manufacturing-heavy states compare to service-oriented urban centers could reveal additional layers of complexity in voter alignment.
+Incorporating regional economic data or policy differences could enhance the explanatory power of future models.
+
+In conclusion, while this paper highlights the significant role of socio-economic factors in shaping political preferences, it also underscores the need for more nuanced and dynamic approaches to understanding voter behavior.
+By addressing these limitations and pursuing the outlined future directions, researchers can build a more comprehensive and context-sensitive picture of the socio-political landscape.
 

### 2024-12-04 scripts/06-model_validation.R

@@ -1,0 +1,23 @@
+#### Preamble ####
+# Purpose: Models... [...UPDATE THIS...]
+# Author: Rohan Alexander [...UPDATE THIS...]
+# Date: 11 February 2023 [...UPDATE THIS...]
+# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# License: MIT
+# Pre-requisites: [...UPDATE THIS...]
+# Any other information needed? [...UPDATE THIS...]
+
+
+#### Workspace setup ####
+library(tidyverse)
+library(rstanarm)
+
+political_preferences <- readRDS(file ="models/political_preferences.rds")
+# Train-test split
+set.seed(853)
+train_indices <- sample(1:nrow(ces_data_reduced), 0.7 * nrow(ces_data_reduced))
+train_data <- ces_data_reduced[train_indices, ]
+test_data <- ces_data_reduced[-train_indices, ]
+
+# Predictions on test data
+test_predictions <- posterior_predict(political_preferences, newdata = test_data)

### 2024-12-04 paper/paper.qmd
appendix, discussion, fixes on prose

@@
 ---
-title: "Younger age groups, women, and lower levels of education are more likely to vote Democractic"
-subtitle: "A logistic regression analysis of demographics in the 2020 US Presidential Election"
+title: "Lower tax brackets and employment levels, and people living in populous areas are more likely to vote for Biden"
 author: 
@@
 date-format: long
-abstract: "In this study, I investigate demographic patterns which influence political support for the US 2020 election, specifically focusing on whether demographics of people tend to vote for Joe Biden or Donald Trump.
-I use a logistic regression model with the variables age, gender, and education to predict which categories in each demographic have the highest chance of voting for Joe Biden.
-I find that the model accurately predicts the trend of younger people, women, and those with the lowest level of education being more likely to support Joe Biden."
+abstract: "In this study, I investigate demographic patterns that influence political support during the 2020 U.S.
+Presidential Election, focusing on whether socio-economic factors like employment status, income level, and living environment influence voters' likelihood to support Joe Biden or Donald Trump.
+I employ a logistic regression model to analyze these predictors and identify trends within each demographic category.
+The results show that urban residents, lower-income groups, and individuals in specific employment categories (e.g., students and retirees) are more likely to support Joe Biden.
+These findings align with broader trends in U.S.
+politics, where socio-economic conditions and regional contexts play pivotal roles in shaping political preferences."
 format: pdf
@@
 # Introduction
-
-The 2020 United States presidential election was a landmark event that encapsulated a confluence of crises and societal shifts.
-Beyond the COVID-19 pandemic, which reshaped healthcare priorities and exposed systemic vulnerabilities, the election underscored debates about economic inequality, particularly as millions faced unemployment during the economic downturn [@Blustein2020; @BLS2020].
-Climate change emerged as a defining issue, with voters increasingly prioritizing environmental policies in response to intensifying natural disasters [@IPCC2020].
-
-Demographic shifts also played a critical role.
-The increasing influence of younger, more diverse voters highlighted generational and racial divides in political priorities [@Frey2020].
-Meanwhile, changes in suburban voting patterns, particularly among college-educated women, illustrated the evolving dynamics of key swing areas [@Cook2020].
-
-Another major factor was the unprecedented role of mail-in voting, which became a focal point for controversy amidst the pandemic and claims of election fraud.
-This shift in voting methods not only increased accessibility for millions but also fueled political and legal battles over election integrity [@NCSL2020].
-Finally, the election was a litmus test for the resilience of democratic institutions in the face of misinformation, political polarization, and challenges to the legitimacy of results [@BenGurion2020; @Sunstein2020].
-
-Analyzing how living environments, employment status, and income levels correlate with political alignment in 2020 offers deep insights into societal trends.
-Research shows that urban residents tend to favor liberal candidates, reflecting the diversity and progressive social policies often prioritized in cities [@bishop2019bigsort].
-Conversely, rural voters strongly supported Donald Trump, consistent with longstanding trends emphasizing traditional values and conservative policies [@cramer2016politics].
-
-Employment status also played a key role.
-Data from the Cooperative Congressional Election Study (CCES, 2020) revealed that students and retirees, often focused on issues like education and healthcare, leaned more toward Joe Biden, while full-time workers, prioritizing economic stability, showed mixed preferences based on industry and region [@cces2020].
-
-Income disparities further highlighted political divides.
-Lower-income groups, grappling with economic inequality, largely supported Biden due to his policies on healthcare and social safety nets, as noted in Pew Research’s 2020 voter analysis.
-Higher-income voters, on the other hand, leaned Republican, particularly in rural and suburban settings, reflecting preferences for tax policies that align with their financial interests [@pew2020voteranalysis].
-
-These findings display the imapct of personal circumstances and political preferences, mapping not only the political landscape of 2020 but also providing a deeper understanding of the evolving priorities shaping voter behavior.
-
-In this study, I use a logistic regression model to predict which candidate—Joe Biden or Donald Trump—was more likely to receive votes based on factors such as living environment, employment status, and income levels.
-Logistic regression is particularly suited for this analysis as it models binary outcomes, allowing us to estimate the likelihood of a voter aligning with one candidate over the other.
-By focusing on these explanatory variables, the study provides valuable insights into how socioeconomic and regional demographics influenced voter preferences during one of the most polarized elections in modern history.
-
-The primary dataset for this study is the 2020 Cooperative Congressional Election Study (CCES), a stratified sample survey administered by YouGov and sourced from the Harvard Dataverse [@cces2020].
-
-Complementary insights are drawn from Pew Research Center’s 2020 voter analysis, which contextualizes demographic trends across key groups, and the Census Bureau’s Current Population Survey Voting and Registration Supplement, which provides aggregate statistics on voter turnout and demographic characteristics [@pew2020; @CPS2020].
-These additional datasets validate findings from the CCES and enhance the robustness of our conclusions.
-
-My report is structured into five main sections and includes an appendix.
-After the introduction, the first section describes the CCES 2020 dataset, highlighting key attributes and presenting graphs of demographic distributions.
-The second section details the logistic regression model, its rationale, and preliminary findings.
-The third section analyzes how living environment, employment status, and income influenced election outcomes using visualizations and numerical results.
-The fourth section discusses the implications of the findings, potential weaknesses, and future research directions.
-
-The appendix provides an in-depth exploration of survey methodologies, focusing on sampling and observational data related to the CCES 2020 dataset.
-It includes detailed discussions of idealized methodologies, literature comparisons, and simulations to illustrate sampling variability and bias, expanding on points in the Data section.
-
-This analysis was conducted in R, utilizing several packages to streamline data handling, visualization, and modeling.
-Key tools include tidyverse for data manipulation and visualization, dataverse for accessing the CCES dataset, knitr for generating the report, modelsummary for interpreting models, and rstanarm for Bayesian regression analysis [@citeR; @Dataverse; @Knitr; @Modelsummary; @Rstanarm; @Tidybayes; @Dplyr].
-Additionally, ggplot visualizations and elements of the "Discussion" section were developed with the assistance of ChatGPT4 @chatgpt.
+The 2020 United States presidential election highlighted key societal shifts amidst crises.
+Beyond the COVID-19 pandemic, which reshaped healthcare and exposed systemic vulnerabilities, the election underscored debates about economic inequality as unemployment surged [@Blustein2020; @BLS2020].
+Climate change became a priority for many voters, driven by the growing frequency of natural disasters [@IPCC2020].
+
+Demographic patterns also shaped the election.
+Younger, more diverse voters emphasized generational and racial divides in priorities [@Frey2020].
+Suburban voting patterns, particularly among college-educated women, had shifts in key swing areas [@Cook2020].
+The widespread adoption of mail-in voting during the pandemic increased accessibility but fueled political and legal debates over election integrity [@NCSL2020].
+The election also tested democratic institutions against misinformation and polarization [@BenGurion2020; @Sunstein2020].
+
+Living environments strongly influenced political alignment.
+Urban residents predominantly supported Joe Biden, reflecting progressive policy preferences often found in cities [@bishop2019bigsort].
+Conversely, rural voters favored Donald Trump, consistent with support for conservative values [@cramer2016politics].
+
+Employment status was another factor.
+Data from the Cooperative Congressional Election Study (CCES, 2020) indicated that students and retirees leaned toward Biden, likely due to policy interests in education and healthcare.
+Full-time workers exhibited mixed preferences influenced by industry and region [@cces2020].
+
+Income disparities also defined political divides.
+Lower-income groups largely supported Biden, aligning with Democratic policies on healthcare and social safety nets, while higher-income voters leaned Republican, especially in rural and suburban settings, reflecting preferences for tax policies benefiting their financial interests [@pew2020voteranalysis].
+
+This study uses a logistic regression model to predict voting preferences based on living environment, employment status, and income levels.
+Logistic regression is suited for binary outcomes, providing probabilities for voter alignment with each candidate.
+
+The primary dataset is the 2020 Cooperative Congressional Election Study (CCES), a stratified survey conducted by YouGov and sourced from the Harvard Dataverse [@cces2020].
+Supporting data includes Pew Research Center’s 2020 voter analysis and the Census Bureau’s Current Population Survey Voting and Registration Supplement, which validate and contextualize findings [@pew2020; @CPS2020].
+
+The report includes five sections and an appendix.
+After the introduction, the first section describes the CCES dataset, highlighting demographic attributes and distributions.
+The second section explains the logistic regression model and presents preliminary findings.
+The third section analyzes the influence of living environment, employment status, and income on election outcomes using visualizations and numerical results.
+The fourth section discusses the implications, limitations, and directions for future research.
+
+The appendix explores survey methodologies, focusing on sampling and observational data in the CCES dataset.
+It compares methodologies, discusses sampling variability and bias, and uses simulations to expand on points in the Data section.
+
+This analysis was done in R, with the packages tidyverse for data manipulation, dataverse for dataset access, knitr for report generation, modelsummary for model interpretation, and rstanarm for Bayesian regression [@citeR; @Dataverse; @Knitr; @Modelsummary; @Rstanarm; @Tidybayes; @Dplyr].
+Plots and graphs were created with ggplot, and portions of the discussion, data, and appendix were assisted by ChatGPT4 [@chatgpt].
 \newpage
@@
 The dataset for this analysis is the 2020 Cooperative Election Study (CCES) [@ces2020], accessed via the Harvard Dataverse.
-The 2020 iteration surveyed 61,000 respondents, capturing a broad spectrum of U.S.
+The 2020 iteration surveyed 61,000 respondents, capturing a broad range of U.S.
 political opinions.
@@
 
-Sources for @fig-respondents below: [@ANES2020],[@Pew2020],[], [@ces2020].
+Sources for @fig-respondents below: [@ANES2020],[@Pew2020],[@Edison2020], [@ces2020].
 
@@
 The CCES collected data in two waves, immediately before and after the 2020 election.
-This helps capture the shifts in voter attitudes and preferences during the critical final stages of the election campaign [@ces2020].
-Pew Research and Gallup each conduct multiple polls of much smaller scale leading up to the election, which will give a snapshot in time of the voter sentiment but does not capture as broad of a perspective as the CES survey [@Pew2020], [@Gallup2020].
+This helps capture the shifts in voter attitudes and preferences during the final stages of the election campaign [@ces2020].
+Pew Research and Gallup each conduct multiple polls of much smaller scale leading up to the election, which will give a snapshot in time of the voter sentiment but does not capture a longitudinal perspective similarly to the CES survey [@Pew2020].
 
@@
 ## Variables of Interest
-The selection of living environment, employment status, and income level as predictor variables for analyzing political preferences in the 2020 CCES data is guided by prior research that underscores their influence on voting behavior.
+The selection of living environment, employment status, and income level as predictor variables for analyzing political preferences in the 2020 CCES data is guided by prior research that shows their influence on voting behavior.
 
@@
 
-However, this trend is nuanced by education levels and geographic factors.
+However, this trend is affected by education levels and geographic factors.
 High-income earners in urban areas often prioritize social liberalism and climate change policies, aligning with Democratic platforms, while rural high-income earners focus more on fiscal conservatism [@Pew2020; @Edsall2020].
@@
 In the ">500k" category, Trump garners more support, consistent with Republican tax policies favoring high-income earners.
-These patterns underscore the nuanced relationships between employment, income, and political preferences, highlighting the diversity of economic profiles within voter bases.
+These patterns underscore the relationships between employment, income, and political preferences, highlighting the diversity of economic profiles within voter bases.
 
@@
 
-The @fig-heatmap-employment-living provides insights into the distribution of employment statuses across different living environments:
+The @fig-heatmap-employment-living shows the distribution of employment statuses across different living environments:
 
@@
 #| message: false
+
+#### Generate income_summary ####
+income_summary <- cleaned_data %>%
+  group_by(income) %>%
+  summarize(
+    proportion_biden = mean(voted_for == "Biden", na.rm = TRUE),
+    total_count = n()
+  )
 
@@
 ```
-@fig-incomevotes shows that voting for Biden does not seem to have a correlation with voting for Biden.
+
+@fig-incomevotes shows that income level does not seem to have a correlation with voting for Biden.
 
@@
 
-The model is implemented using R, leveraging the rstanarm package for Bayesian logistic regression.
+The model is implemented using R, using the rstanarm package for Bayesian logistic regression.
 The data was preprocessed using the tidyverse package.
@@
 
-Logistic regression is well-suited for categorical and binary-dependent variables, and the explanatory variables in this study are categorical, making it an appropriate choice [@Hosmer2013].
 Employment, income, and living environment are socio-economic factors that significantly influence voting preferences.
@@
 The model avoids unnecessary complexity by focusing on these key variables while maintaining interpretability.
-Priors ($\beta_0, \beta_k \sim \text{Normal}(0, 2.5)$) are weakly informative, ensuring coefficients are regularized without unduly constraining the model [@Gelman2008].
+Priors ($\beta_0, \beta_k \sim \text{Normal}(0, 2.5)$) are weakly informative, ensuring coefficients are regularized without constraining the model [@Gelman2008].
 
@@
 Underlying assumptions include the linearity of predictors on the log-odds scale, independence of observations, and the assumption that predictors adequately capture variability in the data.
-Limitations include the categorical nature of the predictors, which may lose finer granularity, and the potential omission of confounders such as race or education level.
-The model may be less appropriate in cases where relationships are highly non-linear or involve strong interaction effects.
+Limitations include the categorical nature of the predictors, which may lose granularity, and the potential omission of confounders such as race or education level.
+The model may be less appropriate in cases where relationships are non-linear or involve strong interaction effects.
 Alternatives considered include random forests, which are more flexible but lack interpretability, and multinomial logistic regression, which was unnecessary for this binary outcome.
@@
 The validation process includes a train-test split to ensure generalization to new data, ROC Curve and AUC to demonstrate the model's discrimination ability with a moderate AUC value, posterior predictive checks to validate that simulated outcomes align well with observed data, and residual analysis to confirm model assumptions and identify any misfit.
-These steps confirm the robustness and reliability of the logistic regression model for political preferences.
+These steps confirm the reliability of the logistic regression model for this use case.
 
@@
 
-The analysis reveals clear patterns in the relationship between socio-economic predictors and voting preferences in the 2020 U.S.
-Presidential Election.
-Employment, income, and living environment significantly influence political alignment.
-The coefficients in @tbl-logit_summary display the impact of each predictor on the likelihood of voting for Joe Biden versus Donald Trump.
+The resuls of this analysis show that employment, income, and living environment significantly influence political alignment.
+The coefficients in @tbl-model_summary display the impact of each predictor on the likelihood of voting for Joe Biden versus Donald Trump.
 
@@
 #| message: false
+political_preferences <- readRDS(file ="../models/political_preferences.rds")
 
@@
 Predictors with intervals that do not overlap zero indicate significant effects on voting preferences.
-For example, Living Environment: Rural exhibits a significant negative association with voting for Biden, while Income: <10k shows a strong positive effect.
+For example, livingRural area exhibits a significant negative association with voting for Biden, while employmentStudent shows a strong positive effect.
 
@@
 
-```{r}
+```{r out.height="20%"}
 #| label: fig-income-overlay
-#| fig-cap: "Comparison of Predicted and Actual Support for Biden by Income Level"
+#| fig-cap: "Model Prediction vs. Actual Votes by Income Level"
 #| echo: false
@@
 
-```{r}
+```{r out.height="20%"}
 #| label: fig-employment-overlay
-#| fig-cap: "Comparison of Predicted and Actual Support for Biden by Employment"
+#| fig-cap: "Model Prediction vs. Actual Votes by Employment"
 #| echo: false
@@
 
-```{r}
+```{r out.height="20%"}
 #| label: fig-living-overlay
-#| fig-cap: "Overlay of Predicted Odds and Respondent Counts by Living Area"
+#| fig-cap: "Model Prediction vs. Actual Votes by Living Area"
 #| echo: false
@@
 #### Load model ####
-political_preferences <- readRDS("../models/political_preferences.rds")
 
@@
 @fig-living-overlay shows the model has completely different predictions for the living areas, when employment is set to "Full-time" and income is in the "200-500k" range.
-## Model Justification
+
+## Model Evaluation
 
@@
 
-political_preferences <- readRDS(file ="../models/political_preferences.rds")
 # Train-test split
@@
 In @fig-pp, The observed data $y$ is represented by the solid curve.
-The replicated data $y_rep$ is represented by the shaded region, which encompasses the range of simulations based on the posterior distribution of the model parameters.
+The replicated data $y_rep$ is represented by the shaded region, which covers the range of simulations based on the posterior distribution of the model parameters.
 
@@
 The model's predictors were carefully chosen to reflect significant socio-economic dimensions that shape voter behavior.
-Through robust statistical analysis, this study identified meaningful correlations between these predictors and political alignment, shedding light on broader socio-political dynamics in the United States.
-
-The statistical approach balanced complexity and interpretability, leveraging Bayesian regularization to mitigate overfitting while allowing for nuanced insights.
+Through statistical analysis, this study identified meaningful correlations between these predictors and political alignment, shedding light on broader socio-political dynamics in the United States.
+
 By employing a logistic regression model, this paper directly modeled the binary nature of the voting outcome while retaining a clear interpretation of how each predictor contributed to the likelihood of voting for a particular candidate.
@@
 
-## Insights
-
-One key insight from this analysis is the continued significance of the urban-rural divide in American politics.
+## Takeaways
+
+This analysis shows the continued significance of the urban-rural divide in American politics.
 Urban voters were far more likely to support Joe Biden, a trend consistent with existing literature on urban liberalism and rural conservatism [@Pew2020u].
@@
 
-Another critical takeaway is the nuanced role of income in shaping political preferences.
+Another takeaway is the role of income in shaping political preferences.
 Lower-income groups were more likely to support Biden, aligning with the Democratic platform's emphasis on social welfare programs, affordable healthcare, and economic redistribution.
@@
 
-While this study provides valuable insights, several limitations warrant discussion.
+While the model was well-suited for the purpose of predicting binary outcomes, there were several limitations.
 First, the logistic regression model assumes linearity in the log-odds of the predictors, which may oversimplify complex interactions between socio-economic factors and voting behavior.
@@
 The categorical representation of predictors, while simplifying interpretation, also introduces granularity constraints.
-For example, income categories like "50-100k" encompass a broad range of experiences that may differ substantially in their political implications.
-Similarly, the employment categories do not account for nuances such as industry type or job satisfaction, which could have significant effects on political alignment.
+For example, income categories like "50-100k" cover a broad range of experiences that may differ substantially in their political implications.
+Similarly, the employment categories do not account for factors such as industry type or job satisfaction, which could have significant effects on political alignment.
 
@@
 For instance, urban areas in different states may have distinct political leanings influenced by local economic conditions, cultural dynamics, or historical contexts.
-Ignoring these regional nuances may obscure important variations in the data.
+Ignoring these regional differences may obscure important variations in the data.
 
@@
 The findings of this paper open several avenues for future research.
-A key area for exploration is the temporal dimension of voting behavior.
-Incorporating longitudinal data could provide insights into how socio-economic factors interact with political events, economic cycles, or social movements to shape voting preferences over time.
+A key area for discussion is the role of timing in voting behavior.
+Incorporating longitudinal data could describe how socio-economic factors interact with political events, economic cycles, or social movements to shape voting preferences over time.
 Such an approach would allow researchers to examine causal relationships rather than mere associations.
@@
 For example, how do the political preferences of low-income individuals differ by gender or race?
-Addressing these questions would require a more granular dataset and advanced modeling techniques, such as hierarchical or multi-level models [@Highton2009].
+Addressing these questions would require a more granular dataset and modeling techniques such as hierarchical or multi-level models [@Highton2009].
 
@@
 
-Finally, regional analyses could shed light on how local contexts mediate the relationship between socio-economic factors and voting behavior.
-For example, exploring how employment trends in manufacturing-heavy states compare to service-oriented urban centers could reveal additional layers of complexity in voter alignment.
-Incorporating regional economic data or policy differences could enhance the explanatory power of future models.
-
-In conclusion, while this paper highlights the significant role of socio-economic factors in shaping political preferences, it also underscores the need for more nuanced and dynamic approaches to understanding voter behavior.
+Regional analyses are necessary to show how local contexts mediate the relationship between socio-economic factors and voting behavior.
+For example, employment trends in manufacturing-heavy states compared to service-oriented urban centers might cause differences in voter alignment.
+Incorporating regional economic data or policy differences could lead to mroe accurate explanations in future models.
+
+In conclusion, while this paper highlights the significant role of socio-economic factors in shaping political preferences, it also shows the need for more dynamic approaches to understanding voter behavior.
 By addressing these limitations and pursuing the outlined future directions, researchers can build a more comprehensive and context-sensitive picture of the socio-political landscape.
 
+# Appendix: Survey and Sampling Methodology
+
+Overview of Survey Design in the CCES 2020 Dataset
+
+The 2020 Cooperative Election Study (CCES) is a large-scale, national survey designed to capture U.S.
+political behavior and attitudes.
+It employs a stratified sample, drawing from a diverse pool of respondents to achieve national representation The survey design is a collaborative effort between academic institutions and professional survey firms [@Ansolabehere2012].
+
+The CCES includes two waves: a pre-election wave conducted in October 2020 and a post-election wave conducted in November 2020.
+This dual-wave structure allows researchers to capture shifts in voter sentiment and behavior during the election period.
+The survey uses online data collection, a method that benefits from scalability and cost-effectiveness.
+
+## Sampling Techniques
+
+### Stratified Sampling
+
+Stratified sampling ensures that subgroups of the population, such as those defined by age, race, income, or geographic region, are adequately represented.
+For the CCES, stratification is based on demographic characteristics derived from the U.S.
+Census Bureau and voter registration databases.
+This method minimizes sampling error and allows for more precise estimates of population parameters [@Lohr2021].
+
+### Sample Matching
+
+Sample matching is a two-step process used to improve the representativeness of the survey sample.
+First, a target sample is generated based on demographic benchmarks.
+Next, survey respondents are selected from a larger pool to match the characteristics of the target sample.
+The CCES utilizes voter registration data to match respondents, enhancing the validity of findings [@Rivers2006].
+
+### Weighting
+
+Post-stratification weights are applied to the data to correct for any deviations between the survey sample and the target population.
+These weights adjust for factors such as response bias and demographic discrepancies.
+The weights are calculated using iterative proportional fitting, ensuring alignment with key benchmarks like age, gender, race, and geographic distribution [@Gelman2007].
+
+## Observational Data and Challenges
+
+### Strengths of Observational Surveys
+
+Observational surveys like the CCES provide a snapshot of real-world voter behavior, capturing smaller details that experimental designs may overlook.
+They allow for large-scale data collection across diverse populations, which is important for studying complex phenomena like political alignment.
+
+### Limitations
+
+Despite their strengths, observational surveys face challenges such as:
+
+1. Nonresponse Bias: Certain demographic groups may be less likely to participate, leading to underrepresentation.
+
+2. Self-Reporting Bias: Respondents may misreport their voting behavior or demographic characteristics, often due to social desirability [@Ansolabehere2012].
+
+3. Cross-Sectional Nature: The CCES captures data at specific points in time, limiting its ability to account for temporal changes in voter behavior.
+
+## Comparisons with Other Surveys
+
+### American National Election Studies (ANES)
+
+The ANES employs a mixed-mode approach, combining face-to-face interviews with online surveys.
+While ANES provides rich contextual data, its smaller sample size limits the granularity of subgroup analysis compared to the CCES [@Berinsky2014].
+
+### Pew Research Center Surveys
+
+Pew Research specializes in tracking public opinion trends but focuses on shorter questionnaires.
+The CCES’s larger number of more specific questions is more suited for modeling complex relationships like those explored in this paper [@Pew2020].
+
+## Methodological Best Practices
+
+Based on the strengths and limitations of the CCES and other surveys, several best practices emerge:
+
+1. Hybrid Sampling Methods: Combining stratified sampling with random digit dialing or online panel recruitment can improve representativeness.
+
+2. Enhanced Weighting Techniques: Incorporating machine learning algorithms into weighting procedures could reduce bias from hard-to-reach populations.
+
+3. Longitudinal Design: Adding a longitudinal component to surveys like the CCES would allow researchers to track changes in voter behavior over time.
+
+## Future Directions
+
+Future iterations of surveys like the CCES could benefit from integrating administrative data (e.g., tax records or employment data) to validate self-reported responses.
+Additionally, adopting adaptive sampling techniques that dynamically adjust to underrepresented groups could further enhance data quality.
+Finally, the incorporation of qualitative components, such as open-ended responses, could provide richer context to quantitative findings.
+
+By adhering to these methodological enhancements, surveys can accurately represent complex socio-political phenomena.
 \newpage

### 2024-12-04 scripts/00-simulate_data.R

@@ -1,52 +1,51 @@
 #### Preamble ####
-# Purpose: Simulates a dataset of Australian electoral divisions, including the 
-  #state and party that won each division.
-# Author: Rohan Alexander
-# Date: 26 September 2024
-# Contact: rohan.alexander@utoronto.ca
-# License: MIT
-# Pre-requisites: The `tidyverse` package must be installed
-# Any other information needed? Make sure you are in the `starter_folder` rproj
-
+# Purpose: Simulates the 2020 presidential election results by employment, income, and living environment
+# Author: Victor Ma
+# Date: 3 December 2024
+# Contact: victo.ma@mail.utoronto.ca
+# In this simulation I use code from Rohan Alexander
 
 #### Workspace setup ####
 library(tidyverse)
-set.seed(853)
-
 
 #### Simulate data ####
-# State names
-states <- c(
-  "New South Wales",
-  "Victoria",
-  "Queensland",
-  "South Australia",
-  "Western Australia",
-  "Tasmania",
-  "Northern Territory",
-  "Australian Capital Territory"
-)
+set.seed(853)
 
-# Political parties
-parties <- c("Labor", "Liberal", "Greens", "National", "Other")
+num_obs <- 1000
 
-# Create a dataset by randomly assigning states and parties to divisions
-analysis_data <- tibble(
-  division = paste("Division", 1:151),  # Add "Division" to make it a character
-  state = sample(
-    states,
-    size = 151,
-    replace = TRUE,
-    prob = c(0.25, 0.25, 0.15, 0.1, 0.1, 0.1, 0.025, 0.025) # Rough state population distribution
-  ),
-  party = sample(
-    parties,
-    size = 151,
-    replace = TRUE,
-    prob = c(0.40, 0.40, 0.05, 0.1, 0.05) # Rough party distribution
-  )
-)
+us_political_preferences <- tibble( # simulate an example election
+  employment = sample(0:7, size = num_obs, replace = TRUE), # employment categories
+  income = sample(0:4, size = num_obs, replace = TRUE),     # income levels
+  living = sample(0:3, size = num_obs, replace = TRUE),     # living environments
+  support_prob = ((employment + income + living) / 12),    # simplified probability model
+) |>
+  mutate(
+    supports_biden = if_else(runif(n = num_obs) < support_prob, "yes", "no"),
+    employment = case_when(
[…]

### 2024-12-04 scripts/01-test_simulated_data.R

@@ -1,89 +1,141 @@
 #### Preamble ####
-# Purpose: Tests the structure and validity of the simulated Australian 
-  #electoral divisions dataset.
-# Author: Rohan Alexander
-# Date: 26 September 2024
-# Contact: rohan.alexander@utoronto.ca
+# Purpose: Tests the structure and validity of the simulated U.S. political preferences dataset
+# Author: Victor Ma
+# Date: 3 December 2024
+# Contact: victo.ma@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: 
-  # - The `tidyverse` package must be installed and loaded
-  # - 00-simulate_data.R must have been run
-# Any other information needed? Make sure you are in the `starter_folder` rproj
-
+# Pre-requisites:
+# - The `tidyverse` package must be installed and loaded
+# - The simulated dataset must be generated and saved to "data/00-simulated_data/simulated_data.csv"
 
 #### Workspace setup ####
 library(tidyverse)
 
-analysis_data <- read_csv("data/00-simulated_data/simulated_data.csv")
+simulated_data <- read_csv("data/00-simulated_data/simulated_data.csv")
 
-# Test if the data was successfully loaded
-if (exists("analysis_data")) {
+#### Test if the dataset is loaded correctly ####
+if (exists("simulated_data")) {
   message("Test Passed: The dataset was successfully loaded.")
 } else {
   stop("Test Failed: The dataset could not be loaded.")
 }
 
+#### Test data structure ####
 
-#### Test data ####
-
-# Check if the dataset has 151 rows
-if (nrow(analysis_data) == 151) {
-  message("Test Passed: The dataset has 151 rows.")
+# Check if the dataset has the correct number of rows
+expected_rows <- 1000
+if (nrow(simulated_data) == expected_rows) {
+  message("Test Passed: The dataset has ", expected_rows, " rows.")
 } else {
-  stop("Test Failed: The dataset does not have 151 rows.")
+  stop("Test Failed: The dataset does not have ", expected_rows, " rows.")
 }
 
-# Check if the dataset has 3 columns
-if (ncol(analysis_data) == 3) {
-  message("Test Passed: The dataset has 3 columns.")
+# Check if the dataset has the correct number of columns
+expected_columns <- 4
+if (ncol(simulated_data) == expected_columns) {
+  message("Test Passed: The dataset has ", expected_columns, " columns.")
 } else {
-  stop("Test Failed: The dataset does not have 3 columns.")
[…]

### 2024-12-04 scripts/03-clean_data.R

@@ -1,7 +1,7 @@
 #### Preamble ####
 # Purpose: Cleans the raw 2020 CCES data recorded by Schaffner, Brian et.al 
 # Author: Victor Ma
-# Date: 03 December 2024
+# Date: 3 December 2024
 # Contact: victo.ma@mail.utoronto.ca
 
 #### Workspace setup ####

### 2024-12-04 scripts/04-test_analysis_data.R

@@ -1,69 +1,123 @@
 #### Preamble ####
-# Purpose: Tests... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 26 September 2024 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Tests the structure and validity of the cleaned CCES 2020 dataset
+# Author: Victor Ma
+# Date: 3 December 2024
+# Contact: victo.ma@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
-
+# Pre-requisites:
+# - The `tidyverse` and `testthat` packages must be installed and loaded
 
 #### Workspace setup ####
 library(tidyverse)
 library(testthat)
 
-data <- read_csv("data/02-analysis_data/analysis_data.csv")
+cleaned_data <- read_csv("../data/02-analysis_data/analysis_data.csv")
 
+#### Test data structure ####
 
-#### Test data ####
-# Test that the dataset has 151 rows - there are 151 divisions in Australia
-test_that("dataset has 151 rows", {
-  expect_equal(nrow(analysis_data), 151)
+# Test that the dataset has rows
+test_that("dataset has rows", {
+  expect_gt(nrow(cleaned_data), 0)
 })
 
-# Test that the dataset has 3 columns
-test_that("dataset has 3 columns", {
-  expect_equal(ncol(analysis_data), 3)
+# Test that the dataset has exactly 4 columns
+test_that("dataset has 4 columns", {
+  expect_equal(ncol(cleaned_data), 4)
 })
 
-# Test that the 'division' column is character type
-test_that("'division' is character", {
-  expect_type(analysis_data$division, "character")
+# Test that the column names are correct
+expected_columns <- c("voted_for", "living", "employment", "income")
+test_that("column names are correct", {
+  expect_equal(names(cleaned_data), expected_columns)
 })
 
-# Test that the 'party' column is character type
-test_that("'party' is character", {
-  expect_type(analysis_data$party, "character")
+#### Test data types ####
+
+# Test that each column has the expected type
+test_that("column data types are correct", {
+  expect_type(cleaned_data$voted_for, "character")
+  expect_type(cleaned_data$living, "character")
+  expect_type(cleaned_data$employment, "character")
+  expect_type(cleaned_data$income, "character")
 })
 
[…]

### 2024-12-04 scripts/06-model_validation.R

@@ -1,23 +1,0 @@
-#### Preamble ####
-# Purpose: Models... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
-# License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
-
-
-#### Workspace setup ####
-library(tidyverse)
-library(rstanarm)
-
-political_preferences <- readRDS(file ="models/political_preferences.rds")
-# Train-test split
-set.seed(853)
-train_indices <- sample(1:nrow(ces_data_reduced), 0.7 * nrow(ces_data_reduced))
-train_data <- ces_data_reduced[train_indices, ]
-test_data <- ces_data_reduced[-train_indices, ]
-
-# Predictions on test data
-test_predictions <- posterior_predict(political_preferences, newdata = test_data)

### 2024-12-15 paper/paper.qmd
update prose

@@
 # Introduction
-The 2020 United States presidential election highlighted key societal shifts amidst crises.
-Beyond the COVID-19 pandemic, which reshaped healthcare and exposed systemic vulnerabilities, the election underscored debates about economic inequality as unemployment surged [@Blustein2020; @BLS2020].
-Climate change became a priority for many voters, driven by the growing frequency of natural disasters [@IPCC2020].
-
-Demographic patterns also shaped the election.
-Younger, more diverse voters emphasized generational and racial divides in priorities [@Frey2020].
-Suburban voting patterns, particularly among college-educated women, had shifts in key swing areas [@Cook2020].
-The widespread adoption of mail-in voting during the pandemic increased accessibility but fueled political and legal debates over election integrity [@NCSL2020].
-The election also tested democratic institutions against misinformation and polarization [@BenGurion2020; @Sunstein2020].
-
-Living environments strongly influenced political alignment.
-Urban residents predominantly supported Joe Biden, reflecting progressive policy preferences often found in cities [@bishop2019bigsort].
-Conversely, rural voters favored Donald Trump, consistent with support for conservative values [@cramer2016politics].
-
-Employment status was another factor.
-Data from the Cooperative Congressional Election Study (CCES, 2020) indicated that students and retirees leaned toward Biden, likely due to policy interests in education and healthcare.
-Full-time workers exhibited mixed preferences influenced by industry and region [@cces2020].
-
-Income disparities also defined political divides.
-Lower-income groups largely supported Biden, aligning with Democratic policies on healthcare and social safety nets, while higher-income voters leaned Republican, especially in rural and suburban settings, reflecting preferences for tax policies benefiting their financial interests [@pew2020voteranalysis].
-
-This study uses a logistic regression model to predict voting preferences based on living environment, employment status, and income levels.
-Logistic regression is suited for binary outcomes, providing probabilities for voter alignment with each candidate.
-
-The primary dataset is the 2020 Cooperative Congressional Election Study (CCES), a stratified survey conducted by YouGov and sourced from the Harvard Dataverse [@cces2020].
-Supporting data includes Pew Research Center’s 2020 voter analysis and the Census Bureau’s Current Population Survey Voting and Registration Supplement, which validate and contextualize findings [@pew2020; @CPS2020].
+The 2020 United States presidential election highlighted societal shifts during crises.
+The COVID-19 pandemic reshaped healthcare and exposed systemic vulnerabilities, while economic inequality became central as unemployment surged [@Blustein2020; @BLS2020].
+Climate change also emerged as a priority, driven by natural disasters [@IPCC2020].
+
+Demographic patterns influenced the election.
+Younger, diverse voters emphasized generational and racial divides in priorities [@Frey2020].
+Suburban voting patterns among college-educated women revealed shifts in key areas [@Cook2020].
+Mail-in voting increased accessibility during the pandemic but fueled debates over election integrity [@NCSL2020].
+The election tested democratic institutions against misinformation and polarization [@BenGurion2020; @Sunstein2020].
+
+Living environments strongly shaped political alignment.
+Urban residents predominantly supported Joe Biden, reflecting progressive policy preferences [@bishop2019bigsort].
+Rural voters favored Donald Trump, consistent with conservative values [@cramer2016politics].
+Employment status and income further defined political divides.
+Students and retirees leaned toward Biden, driven by concerns about healthcare and education [@cces2020].
+Higher-income voters leaned Republican, especially in rural and suburban areas [@pew2020voteranalysis].
+
+This study estimates the probability of voting for Joe Biden versus Donald Trump, conditional on living environment, employment status, and income levels.
+Logistic regression, suited for binary outcomes, is used to model these relationships.
+The analysis relies on the 2020 Cooperative Congressional Election Study (CCES), a stratified survey conducted by YouGov and sourced from the Harvard Dataverse [@cces2020].
+Supporting data from Pew Research and the Census Bureau’s Current Population Survey validate findings [@pew2020; @CPS2020].
+
+The estimand is the effect of income level, employment status, and living area on the odds of voting for Biden, measured as the change in predicted odds associated with each predictor, holding other factors constant.
 
 The report includes five sections and an appendix.
-After the introduction, the first section describes the CCES dataset, highlighting demographic attributes and distributions.
-The second section explains the logistic regression model and presents preliminary findings.
-The third section analyzes the influence of living environment, employment status, and income on election outcomes using visualizations and numerical results.
-The fourth section discusses the implications, limitations, and directions for future research.
+After the introduction, the first section describes the CCES dataset, highlighting demographic distributions.
+The second section explains the logistic regression model and presents findings.
+The third section analyzes the effects of living environment, employment, and income on voting using visualizations and numerical results.
+The fourth section discusses implications, limitations, and directions for future research.
 
 The appendix explores survey methodologies, focusing on sampling and observational data in the CCES dataset.
-It compares methodologies, discusses sampling variability and bias, and uses simulations to expand on points in the Data section.
-
-This analysis was done in R, with the packages tidyverse for data manipulation, dataverse for dataset access, knitr for report generation, modelsummary for model interpretation, and rstanarm for Bayesian regression [@citeR; @Dataverse; @Knitr; @Modelsummary; @Rstanarm; @Tidybayes; @Dplyr].
-Plots and graphs were created with ggplot, and portions of the discussion, data, and appendix were assisted by ChatGPT4 [@chatgpt].
+It compares methodologies, examines sampling variability and bias, and uses simulations to expand on points in the data section.
+
+This analysis was conducted in R, using tidyverse for data manipulation, dataverse for dataset access, knitr for report generation, modelsummary for model interpretation, and rstanarm for Bayesian regression [@citeR; @Dataverse; @Knitr; @Modelsummary; @Rstanarm; @Tidybayes; @Dplyr].
+Visualizations were created with ggplot, and portions of the discussion and appendix were assisted by ChatGPT4 [@chatgpt].
+
 \newpage
@@
 
-This paper investigated the influence of socio-economic factors—employment status, income level, and living environment—on voting preferences during the 2020 U.S.
+This paper examines how socio-economic factors—employment status, income level, and living environment—shape voting preferences in the 2020 U.S.
 Presidential Election.
-Utilizing the 2020 Cooperative Election Study (CCES) dataset, a logistic regression model was applied to predict support for Joe Biden versus Donald Trump.
-The model incorporated categorical predictors, regularized with Bayesian priors, to account for variability while ensuring interpretability.
-The model's predictors were carefully chosen to reflect significant socio-economic dimensions that shape voter behavior.
-Through statistical analysis, this study identified meaningful correlations between these predictors and political alignment, shedding light on broader socio-political dynamics in the United States.
-
-By employing a logistic regression model, this paper directly modeled the binary nature of the voting outcome while retaining a clear interpretation of how each predictor contributed to the likelihood of voting for a particular candidate.
-The inclusion of categorical variables like employment, income, and living environment provided an avenue to explore how socio-economic realities translate into political preferences.
-
-## Takeaways
-
-This analysis shows the continued significance of the urban-rural divide in American politics.
-Urban voters were far more likely to support Joe Biden, a trend consistent with existing literature on urban liberalism and rural conservatism [@Pew2020u].
-Urban areas, characterized by higher population densities and greater economic and cultural diversity, often prioritize progressive policies such as environmental regulation, healthcare access, and social equity.
-Conversely, rural voters exhibited a marked preference for Donald Trump, reflecting a broader alignment with conservative policies, including lower taxation, gun rights, and traditional social values [@IyengarKrupenkin2018].
-These findings underscore the stark political polarization between urban and rural constituencies, a division that has deepened over successive election cycles.
-
-Another takeaway is the role of income in shaping political preferences.
-Lower-income groups were more likely to support Biden, aligning with the Democratic platform's emphasis on social welfare programs, affordable healthcare, and economic redistribution.
-Middle-income groups displayed more mixed preferences, often reflecting a balance between economic self-interest and policy priorities.
-Interestingly, the wealthiest demographic groups exhibited reduced support for Biden, suggesting that concerns over Democratic tax policies and fiscal regulations may outweigh social considerations in shaping their voting behavior [@Frank2004; @Gelman2009].
-This U-shaped relationship between income and voting patterns highlights the complexity of economic factors in political decision-making.
-
-Employment status further illuminated key dynamics in voter behavior.
-Students and retired individuals showed strong support for Biden, possibly due to generational policy preferences such as student debt relief and Social Security protection.
-Full-time workers demonstrated more balanced voting preferences, reflecting the diversity of this group in terms of socio-economic priorities.
-Notably, individuals in precarious employment situations, such as part-time or temporarily laid-off workers, leaned toward Biden, likely reflecting concerns over job security and healthcare access in the wake of the COVID-19 pandemic.
-
-## Weaknesses
-
-While the model was well-suited for the purpose of predicting binary outcomes, there were several limitations.
-First, the logistic regression model assumes linearity in the log-odds of the predictors, which may oversimplify complex interactions between socio-economic factors and voting behavior.
-Non-linear relationships or interaction effects, such as how income and employment jointly influence voting preferences, may be overlooked in this framework.
-
-The categorical representation of predictors, while simplifying interpretation, also introduces granularity constraints.
-For example, income categories like "50-100k" cover a broad range of experiences that may differ substantially in their political implications.
-Similarly, the employment categories do not account for factors such as industry type or job satisfaction, which could have significant effects on political alignment.
-
-Another limitation lies in the reliance on self-reported survey data.
-Self-reporting is subject to biases such as social desirability and recall inaccuracies, potentially skewing results [@Ansolabehere2012].
-Furthermore, the cross-sectional nature of the CCES dataset captures voter preferences at a single point in time, making it difficult to assess the impact of dynamic political events or campaign strategies on voting behavior [@Gelman2009].
-
-Finally, the model does not account for regional variations within urban and rural categories.
-For instance, urban areas in different states may have distinct political leanings influenced by local economic conditions, cultural dynamics, or historical contexts.
-Ignoring these regional differences may obscure important variations in the data.
-
-## Future Directions
-
-The findings of this paper open several avenues for future research.
-A key area for discussion is the role of timing in voting behavior.
-Incorporating longitudinal data could describe how socio-economic factors interact with political events, economic cycles, or social movements to shape voting preferences over time.
-Such an approach would allow researchers to examine causal relationships rather than mere associations.
-
-Intersectionality is another promising direction.
-Investigating how combinations of factors—such as race, gender, and education—jointly influence voting behavior could yield a richer understanding of political alignment.
-For example, how do the political preferences of low-income individuals differ by gender or race?
-Addressing these questions would require a more granular dataset and modeling techniques such as hierarchical or multi-level models [@Highton2009].
-
-Methodologically, future studies could benefit from experimenting with non-linear models or machine learning approaches.
-These methods can capture complex relationships and interactions that traditional logistic regression might miss.
-For instance, decision tree-based algorithms or neural networks could identify non-obvious patterns in how socio-economic factors influence voting preferences, though at the cost of reduced interpretability.
-
-Regional analyses are necessary to show how local contexts mediate the relationship between socio-economic factors and voting behavior.
-For example, employment trends in manufacturing-heavy states compared to service-oriented urban centers might cause differences in voter alignment.
-Incorporating regional economic data or policy differences could lead to mroe accurate explanations in future models.
-
-In conclusion, while this paper highlights the significant role of socio-economic factors in shaping political preferences, it also shows the need for more dynamic approaches to understanding voter behavior.
-By addressing these limitations and pursuing the outlined future directions, researchers can build a more comprehensive and context-sensitive picture of the socio-political landscape.
+Using the 2020 Cooperative Election Study (CCES) dataset, a logistic regression model was applied to predict support for Joe Biden versus Donald Trump.
+The model incorporated categorical predictors to reflect socio-economic dimensions influencing voter behavior.
+Analysis identified relationships between these predictors and political alignment, highlighting socio-political dynamics in the United States.
+
+The logistic regression model directly addressed the binary nature of the voting outcome, providing interpretable results on how each predictor influenced voting likelihood.
+Including categorical variables such as employment, income, and living environment allowed analysis of how socio-economic realities influence political preferences.
+
+### Takeaways
+
+The urban-rural divide continues to shape American politics.
+Urban voters predominantly supported Joe Biden, aligning with policies related to healthcare, environmental regulation, and equity [@Pew2020u].
+Rural voters favored Donald Trump, supporting policies related to lower taxes, gun rights, and traditional values [@IyengarKrupenkin2018].
+This polarization reflects growing differences between urban and rural constituencies.
+
+Income also influenced voting behavior.
+Lower-income groups supported Biden, consistent with policies prioritizing social welfare and economic redistribution.
+Middle-income groups showed mixed preferences, balancing economic priorities with policy concerns.
+Higher-income voters leaned Republican, likely due to concerns over Democratic tax policies [@Frank2004; @Gelman2009].
+These trends reflect a U-shaped relationship between income and political preferences.
+
+Employment status further influenced voter behavior.
+Students and retirees supported Biden, reflecting priorities such as debt relief and Social Security.
+Full-time workers demonstrated varied preferences due to diverse priorities.
+Part-time and temporarily laid-off workers leaned toward Biden, likely reflecting concerns about job security and healthcare during the COVID-19 pandemic.
+
+### Weaknesses
+
+The model has several limitations.
+Logistic regression assumes linearity in log-odds, potentially oversimplifying relationships between predictors and voting behavior.
+Interaction effects, such as between income and employment, were not fully explored.
+
+Categorical predictors simplify analysis but reduce specificity.
+For instance, income ranges like "50-100k" include diverse experiences that may influence voting behavior differently.
+Employment categories exclude factors such as industry type or job stability, which likely play significant roles in voter preferences.
+
+Reliance on self-reported survey data introduces biases like social desirability and recall errors [@Ansolabehere2012].
+The cross-sectional nature of the CCES dataset limits the analysis to static preferences, ignoring changes over time and campaign effects [@Gelman2009].
+Additionally, regional variations within urban and rural areas are not accounted for, which may obscure important differences.
+
+### Future Directions
+
+Future research could analyze how socio-economic factors interact with political events or social movements using longitudinal data, enabling causal analysis of voting behavior over time.
+Examining intersections of socio-economic factors with race, gender, or education could offer a richer understanding of voter preferences [@Highton2009].
+
+Non-linear models and machine learning methods could uncover relationships missed by logistic regression.
+Decision tree-based models or neural networks could identify patterns in voter behavior, though they may sacrifice interpretability.
+
+Regional analyses are necessary to understand how local economic and cultural contexts influence voting preferences.
+Comparing manufacturing-heavy states with service-oriented urban centers could refine explanations of socio-economic effects on voting.
+
+This paper demonstrates the role of socio-economic factors in shaping political preferences.
+Addressing these limitations and pursuing these directions could deepen understanding of voter behavior and the socio-political landscape.
 
@@
 
-Overview of Survey Design in the CCES 2020 Dataset
-
-The 2020 Cooperative Election Study (CCES) is a large-scale, national survey designed to capture U.S.
+### Overview of Survey Design in the CCES 2020 Dataset
+
+The 2020 Cooperative Election Study (CCES) is a large-scale survey capturing U.S.
 political behavior and attitudes.
-It employs a stratified sample, drawing from a diverse pool of respondents to achieve national representation The survey design is a collaborative effort between academic institutions and professional survey firms [@Ansolabehere2012].
-
-The CCES includes two waves: a pre-election wave conducted in October 2020 and a post-election wave conducted in November 2020.
-This dual-wave structure allows researchers to capture shifts in voter sentiment and behavior during the election period.
-The survey uses online data collection, a method that benefits from scalability and cost-effectiveness.
-
-## Sampling Techniques
-
-### Stratified Sampling
-
-Stratified sampling ensures that subgroups of the population, such as those defined by age, race, income, or geographic region, are adequately represented.
-For the CCES, stratification is based on demographic characteristics derived from the U.S.
+Administered by YouGov with academic oversight, it uses stratified sampling to ensure national representation [@Ansolabehere2012].
+The survey includes a pre-election wave in October 2020 and a post-election wave in November 2020, allowing comparisons of voter attitudes and behaviors during the election period.
+Data collection occurs online, offering scalability and efficiency compared to traditional methods.
+
+### Sampling Techniques
+
+#### Stratified Sampling
+
+Stratified sampling ensures proportional representation of subgroups based on demographic data from the U.S.
 Census Bureau and voter registration databases.
-This method minimizes sampling error and allows for more precise estimates of population parameters [@Lohr2021].
-
-### Sample Matching
-
-Sample matching is a two-step process used to improve the representativeness of the survey sample.
-First, a target sample is generated based on demographic benchmarks.
-Next, survey respondents are selected from a larger pool to match the characteristics of the target sample.
-The CCES utilizes voter registration data to match respondents, enhancing the validity of findings [@Rivers2006].
-
-### Weighting
-
-Post-stratification weights are applied to the data to correct for any deviations between the survey sample and the target population.
-These weights adjust for factors such as response bias and demographic discrepancies.
-The weights are calculated using iterative proportional fitting, ensuring alignment with key benchmarks like age, gender, race, and geographic distribution [@Gelman2007].
-
-## Observational Data and Challenges
-
-### Strengths of Observational Surveys
-
-Observational surveys like the CCES provide a snapshot of real-world voter behavior, capturing smaller details that experimental designs may overlook.
-They allow for large-scale data collection across diverse populations, which is important for studying complex phenomena like political alignment.
-
-### Limitations
-
-Despite their strengths, observational surveys face challenges such as:
-
-1. Nonresponse Bias: Certain demographic groups may be less likely to participate, leading to underrepresentation.
-
-2. Self-Reporting Bias: Respondents may misreport their voting behavior or demographic characteristics, often due to social desirability [@Ansolabehere2012].
-
-3. Cross-Sectional Nature: The CCES captures data at specific points in time, limiting its ability to account for temporal changes in voter behavior.
-
-## Comparisons with Other Surveys
-
-### American National Election Studies (ANES)
-
-The ANES employs a mixed-mode approach, combining face-to-face interviews with online surveys.
-While ANES provides rich contextual data, its smaller sample size limits the granularity of subgroup analysis compared to the CCES [@Berinsky2014].
-
-### Pew Research Center Surveys
-
-Pew Research specializes in tracking public opinion trends but focuses on shorter questionnaires.
-The CCES’s larger number of more specific questions is more suited for modeling complex relationships like those explored in this paper [@Pew2020].
-
-## Methodological Best Practices
-
-Based on the strengths and limitations of the CCES and other surveys, several best practices emerge:
-
-1. Hybrid Sampling Methods: Combining stratified sampling with random digit dialing or online panel recruitment can improve representativeness.
-
-2. Enhanced Weighting Techniques: Incorporating machine learning algorithms into weighting procedures could reduce bias from hard-to-reach populations.
-
-3. Longitudinal Design: Adding a longitudinal component to surveys like the CCES would allow researchers to track changes in voter behavior over time.
-
-## Future Directions
-
-Future iterations of surveys like the CCES could benefit from integrating administrative data (e.g., tax records or employment data) to validate self-reported responses.
-Additionally, adopting adaptive sampling techniques that dynamically adjust to underrepresented groups could further enhance data quality.
-Finally, the incorporation of qualitative components, such as open-ended responses, could provide richer context to quantitative findings.
-
-By adhering to these methodological enhancements, surveys can accurately represent complex socio-political phenomena.
+This approach minimizes sampling variability and improves the precision of population estimates [@Lohr2021].
+Subgroup analyses, such as examining rural voting patterns or income brackets, are possible due to this method.
+
+#### Sample Matching
+
+Sample matching enhances representativeness by selecting respondents from a pool based on predefined demographic benchmarks.
+The CCES matches respondents to a target sample derived from voter registration data, reducing biases associated with convenience sampling [@Rivers2006].
+
+#### Weighting
+
+Post-stratification weights correct deviations between the survey sample and the target population.
+Iterative proportional fitting adjusts for age, gender, race, and geographic imbalances [@Gelman2007].
+Weighting ensures the dataset reflects national trends, accounting for differences in response rates among demographic groups.
+
+### Observational Data and Challenges
+
+#### Strengths of Observational Surveys
+
+Observational surveys like the CCES provide data on real-world voter behavior across diverse populations.
+The CCES’s large sample size, exceeding 61,000 respondents, allows detailed analyses of interactions between income, employment, and geography.
+
+#### Limitations
+
+1. **Nonresponse Bias**: Underrepresentation of certain groups, such as younger or less-educated individuals, skews results.
+2. **Self-Reporting Bias**: Respondents may misreport voting behavior or demographic details due to social desirability or recall errors [@Ansolabehere2012].
+3. **Cross-Sectional Nature**: The CCES captures data at specific points in time, limiting its ability to assess changes over time.
+4. **Internet Access Disparities**: Online surveys may exclude populations with limited internet access, such as older or rural voters.
+
+### Comparisons with Other Surveys
+
+#### American National Election Studies (ANES)
+
+The ANES combines face-to-face interviews with online surveys.
+Its smaller sample size limits subgroup analyses, but its detailed questions on political efficacy and ideology complement the CCES’s broader demographic focus [@Berinsky2014].
+
+#### Pew Research Center Surveys
+
+Pew Research surveys focus on shorter questionnaires tracking public opinion trends.
+While useful for snapshots, the CCES’s larger sample size and detailed questions better support modeling relationships between variables like income, employment, and location [@Pew2020].
+
+### Methodological Best Practices
+
+1. **Hybrid Sampling Methods**: Combining stratified sampling with other methods, such as online panel recruitment, improves representativeness.
+2. **Enhanced Weighting Techniques**: Machine learning in weighting processes can reduce biases among hard-to-reach groups.
+3. **Longitudinal Components**: Tracking individuals over time allows analysis of how voter behavior changes with events.
+4. **Integration of Administrative Data**: Linking survey responses to administrative records, such as tax data, improves accuracy and reliability.
+
+### Future Directions
+
+To address challenges, future surveys like the CCES could adopt innovations such as:
+
+- **Adaptive Sampling**: Adjusting strategies during data collection improves coverage of underrepresented groups.
+- **Qualitative Data**: Including open-ended questions provides deeper understanding of voter motivations.
+- **Non-Linear Models**: Bayesian hierarchical models improve estimation of subgroup effects, particularly for small populations.
+- **Mobile Platforms**: Mobile-friendly surveys increase participation among younger and low-income groups, addressing access disparities.
+
+These enhancements will improve survey accuracy and better capture the complexity of voter behavior.
+
 \newpage

### 2024-12-16 paper/paper.qmd
last changes

@@
 
-The estimand is the effect of income level, employment status, and living area on the odds of voting for Biden, measured as the change in predicted odds associated with each predictor, holding other factors constant.
+The estimand is the probability that a voter supports Joe Biden as opposed to Donald Trump, conditional on the socio-economic factors of living environment, employment status, and income levels.
 
@@
 
-The appendix explores survey methodologies, focusing on sampling and observational data in the CCES dataset.
-It compares methodologies, examines sampling variability and bias, and uses simulations to expand on points in the data section.
+The appendix explores survey methodologies, focusing on sampling and observational data in the CCES dataset. It compares methodologies and discusses sampling variability and biases.
 
@@
 Visualizations were created with ggplot, and portions of the discussion and appendix were assisted by ChatGPT4 [@chatgpt].
-
 \newpage
